---
title: "1.1 — CD4 Integration and Annotation — Cenerenti et al."
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script reproduces the integration and annotation workflow starting from
[scRNA-seq/cache/all_samples_merged.rds](scRNA-seq/cache/all_samples_merged.rds),
which is generated by [scRNA-seq/0.Data_processing.Rmd](scRNA-seq/0.Data_processing.Rmd).

Workflow summary:

1. Load merged Seurat object from script 0
2. Integrate total non-specific CD4 cells with STACAS
3. Perform unsupervised clustering
4. Manually annotate major CD4 states
5. Annotate the same cells using ProjecTILs CD4 reference map
6. Project tumor-specific CD4 cells onto the integrated/manual reference

## Inputs

- `cache/all_samples_merged.rds` from script 0
- ProjecTILs CD4 reference map from figshare:
  <https://figshare.com/ndownloader/files/43794750>

## Outputs

- `cache/cd4_total_integrated_manual.rds`
- `cache/Projectils_prediction_totalCD4.rds`
- `cache/tumor_specific_projection.rds`

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7
)

# Load utility functions for this analysis
source(here::here("scRNA-seq", "utils.R"))
source(here::here("scRNA-seq", "klrg1_ratio_utils.R"))

```

## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(Seurat)
library(STACAS)
library(ProjecTILs)
library(UCell)       # gene-set module scoring
library(tidyverse)
library(patchwork)
library(ggpubr)      # stat_compare_means() for UCell boxplots
library(data.table)  # rbindlist() for ProjecTILs prediction tables
library(here)        # project-relative paths
```

```{r paths}
cache_dir <- here::here("scRNA-seq", "cache")
in_rds    <- file.path(cache_dir, "all_samples_merged.rds")

if (!file.exists(in_rds)) {
  stop("Input not found: ", in_rds,
       "\nPlease run scRNA-seq/0.Data_processing.Rmd first.")
}
```

---

# 1 — Load merged object and harmonize metadata

```{r load-data}
seu <- readRDS(in_rds)
cat("Loaded merged object: ", ncol(seu), " cells x ", nrow(seu), " genes\n", sep = "")
```

```{r metadata-cleanup}
# Keep metadata naming consistent across scripts.
seu@meta.data <- seu@meta.data %>%
  mutate(
    specificity = case_when(
      is.na(specificity) ~ "non_specific",
      grepl("virus|pathogen", specificity, ignore.case = TRUE) ~ "virus_specific",
      TRUE ~ specificity
    ),
    tissue = ifelse(grepl("TIL", tissue, ignore.case = TRUE), "TILN", tissue)
  )

# Convenience label to split by hashtag/capture identity
seu$hashtag <- sapply(seu$MULTI_classification,
                      function(x){
                        strsplit(x, "-")[[1]][1]
                      })

cat("Specificity distribution:\n")
print(table(seu$specificity, useNA = "ifany"))
```

---

# 2 — Subset total (non-specific) CD4 cells

```{r subset-cd4-total}
cd4_total <- subset(seu, subset = specificity == "non_specific" & T_subset == "CD4")
cat("CD4 total (non-specific): ", ncol(cd4_total), " cells\n", sep = "")

print(table(cd4_total$tissue, useNA = "ifany"))
print(table(cd4_total$patient, useNA = "ifany"))
```

---

# 3 — Integration with STACAS

Cells are split by capture and integrated with STACAS.

```{r stacas-integrate}
split_obj <- SplitObject(cd4_total, split.by = "hashtag")

int <- Run.STACAS(
  object.list      = split_obj,
  dims             = 1:30,
  anchor.features  = 1000
)

int <- RunUMAP(int, dims = 1:30, verbose = FALSE)
UMAPPlot(int) + ggtitle("Integrated CD4 total cells")
```

---

# 4 — Unsupervised clustering

```{r clustering}
DefaultAssay(int) <- "integrated"
int <- FindNeighbors(int, dims = 1:30)
int <- FindClusters(int, resolution = 0.4, algorithm = 1)
```

Corresponding to Supp Fig 1B
```{r umap-clustering}
UMAPPlot(int, group.by = "integrated_snn_res.0.4", label = TRUE, repel = TRUE) +
  ggtitle("Unsupervised clusters (res=0.4)")
```

---

# 5 — Marker inspection and manual annotation

## 5.1 Marker panels

```{r marker-panels}
cd4_naive_markers <- c("LEF1", "TCF7", "SELL", "S1PR1", "IL7R", "CCR7")
cd4_memory_markers <- c("CXCR3", "CCL5", "PRDM1", "KLRB1", "GZMK", "ITGB1")
cd4_cytotoxic_markers <- c("GZMB", "GNLY", "NKG7", "FGFBP2", "PRF1", "CST7", "GZMA")
tfh_markers <- c("CXCR5", "PDCD1", "CXCL13", "IL21", "TOX")
treg_markers <- c("FOXP3", "IL2RA", "CTLA4", "TIGIT")

marker_features <- unique(c(
  cd4_naive_markers,
  cd4_memory_markers,
  cd4_cytotoxic_markers,
  tfh_markers,
  treg_markers
))
```

Corresponding to Supp Fig 1C
```{r marker-plots, fig.width=14, fig.height=8}
DefaultAssay(int) <- "RNA"

# Use custom dotplot function for publication-quality markers visualization
dotplot_seurat_style(
  object = int,
  features = marker_features,
  ident_col = "integrated_snn_res.0.4",
  split_by = NULL,
  assay = "RNA"
) +
  ggplot2::ggtitle("Marker expression per unsupervised cluster")
```

## 5.2 Manual annotation

Assign cluster identity based on previously used markers.

```{r manual-annotation}
int@meta.data <- int@meta.data %>% 
  mutate(annotation = factor(integrated_snn_res.0.4,
                             levels = c(0, 2, 3, 1, 4, 5, 6),
                             labels = c(rep("Naive-like", 3),
                                        "Memory", "Cytotoxic",
                                        "Treg", "TFH")),
         annotation = factor(annotation,
                             levels = c("TFH", "Treg", "Naive-like",
                                        "Memory", "Cytotoxic"))
         )

cat("Manual annotation counts:\n")
print(table(int$annotation, useNA = "ifany"))
```

Corresponding to Figure 1D

```{r umap-annotation}
UMAPPlot(int, group.by = "annotation",
         label = TRUE,
         repel = TRUE,
         cols = cd4_annotation_palette) +
  ggtitle("Manual annotation")
```


## 5.3 Custom annotation dotplots

Corresponding to Figure 1E (left)

```{r custom-dotplots, fig.width=12, fig.height=6}
# Joined dotplot by manual annotation
dotplot_seurat_style(
  object = int,
  features = marker_features,
  ident_col = "annotation",
  split_by = NULL,
  assay = "RNA"
) +
  ggplot2::ggtitle("Marker expression per manual annotation")
```
Corresponding to Figure 1E (right)

```{r split-tissue-dotplot-cytotoxic, fig.width=14, fig.height=8}
dotplot_seurat_style(
  object = int,
  features = marker_features,
  subset_col = "annotation",
  subset_values = "Cytotoxic",
  ident_col = "tissue",
  split_by = NULL,
  scale_expression = F
) +
  scale_color_gradient2(high = "#FF7F00",
                        low = "gray80",
                        mid = "gray80",
                        midpoint = 0.7) +
  labs(color = "Average\nExpression")
```

---

# 6 — Reference annotation with ProjecTILs

## 6.1 Download and load CD4 reference map

```{r download-projectils-ref}
ref_url <- "https://ndownloader.figshare.com/files/43794750"
ref_path <- file.path(cache_dir, "CD4T_human_ref_v2.rds")

if (!file.exists(ref_path)) {
  cat("Downloading ProjecTILs CD4 reference map from figshare ...\n")
  download.file(ref_url, destfile = ref_path, mode = "wb", quiet = FALSE)
} else {
  cat("Reference map already present: ", ref_path, "\n", sep = "")
}

cd4ref <- ProjecTILs::load.reference.map(ref_path)
UMAPPlot(cd4ref) + ggtitle("ProjecTILs CD4 reference map")
```

## 6.2 Classify total CD4 cells with ProjecTILs

```{r projectils-classify-total}
pred_list <- list()
# split by capture, supposed to have the batch effect
split_cd4 <- SplitObject(cd4_total, split.by = "MULTI_classification")

for (i in seq_along(split_cd4)) {
  obj_name <- names(split_cd4)[i]
  cat("Processing ", i, "/", length(split_cd4), ": ", obj_name, "\n", sep = "")

  tmp <- ProjecTILs.classifier(
    split_cd4[[i]],
    ref = cd4ref,
    filter.cells = FALSE,
    skip.normalize = FALSE
  )

  pred_list[[i]] <- tmp@meta.data[, "functional.cluster", drop = FALSE]
}

projectils_prediction <- pred_list %>%
  lapply(function(x) {
    x$cell_id <- rownames(x)
    x
  }) %>%
  data.table::rbindlist()

saveRDS(projectils_prediction,
        file.path(cache_dir, "Projectils_prediction_totalCD4.rds"))
```

## 6.3 Compare manual vs reference annotation

Corresponding to Supp Figure 1D

```{r compare-manual-reference, fig.width=10, fig.height=7}
align_annot <- left_join(
  int@meta.data[, "annotation", drop = FALSE] %>% tibble::rownames_to_column("cell_id"),
  projectils_prediction,
  by = "cell_id"
)

align_plot <- align_annot %>%
  filter(!is.na(annotation), !is.na(functional.cluster)) %>%
  count(annotation, functional.cluster, name = "ncells") %>%
  group_by(annotation) %>%
  mutate(rel_pct = ncells / sum(ncells)) %>%
  ungroup() %>%
  ggplot(aes(annotation, rel_pct, fill = functional.cluster)) +
  geom_col(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = cd4ref@misc$atlas.palette) +
  labs(x = "Manual annotation",
       y = "Proportion of reference-map labels",
       fill = "Reference annotation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

align_plot
```

---

# 7 — Projection of tumor-specific CD4 cells

Tumor-specific CD4 cells are projected onto a reference built from the
integrated/manual total CD4 map.

## 7.1 Build projection reference from integrated object

```{r build-manual-reference}
ref_manual <- make.reference(
  ref = int,
  assay = "integrated",
  ndim = 50,
  seed = 22,
  recalculate.umap = FALSE,
  nfeatures = 1500,
  annotation.column = "annotation"
)

DimPlot(ref_manual, group.by = "annotation") + ggtitle("Reference used for projection")
```

## 7.2 Project tumor-specific cells

```{r project-tumor-specific}
tumor_cd4 <- subset(seu, subset = specificity == "tumor_specific" & T_subset == "CD4")
cat("Tumor-specific CD4 cells: ", ncol(tumor_cd4), "\n", sep = "")

proj <- Run.ProjecTILs(
  query = tumor_cd4,
  ref = ref_manual,
  filter.cells = FALSE,
  ndim = 50,
  ncores = 1,
  progressbar = FALSE
)

proj$functional.cluster <- factor(proj$functional.cluster,
                                  levels = levels(int$annotation))
```

Corresponding to Supp Fig F

```{r projection-plot-tumorspecific}
plot.projection(
  ref = ref_manual,
  query = proj,
  linesize = 0,
  pointsize = 1.2,
  cols = cd4_annotation_palette
) + ggtitle("Projection of tumor-specific CD4 cells")
```

```{r projection-summary}
cat("Projected state counts:\n")
print(table(proj$functional.cluster, useNA = "ifany"))

proj_prop <- prop.table(table(proj$functional.cluster))
print(round(proj_prop, 3))
```

## 7.3 Gene markers tumor-specific cells

Corresponding to Figure 1G

```{r custom-dotplots, fig.width=12, fig.height=6}
# Joined dotplot by manual annotation
dotplot_seurat_style(
  object = proj,
  features = marker_features,
  ident_col = "functional.cluster",
  split_by = NULL,
  assay = "RNA"
) +
  ggplot2::ggtitle("Marker expression per annotation")
```

---

# 8 — UCell cytotoxic signature score

We score every cell for the cytotoxic and TFH gene signatures using UCell,
then compare mean scores across tissues within the Cytotoxic annotation state.

## 8.1 Add UCell scores

```{r ucell-scores}
DefaultAssay(int)  <- "RNA"

(ucell_feats <- list(
  Cytotoxic = cd4_cytotoxic_markers
))

int <- AddModuleScore_UCell(int,
                            features = ucell_feats,
                            ncores = 8,
                            assay = "RNA",
                            slot = "counts",
                            name = "")
```

## 8.2 Cytotoxic score — total CD4 (PBMC vs TILN)

Corresponding to Supp Fig 1E

```{r ucell-total}
total_ctl <- int@meta.data %>%
  filter(annotation == "Cytotoxic") %>%
  select(patient, tissue, Cytotoxic) %>%
  group_by(patient, tissue) %>%
  summarize(mean_ctl = mean(Cytotoxic), n_cells = n(), .groups = "drop") %>%
  filter(n_cells > 5)

set.seed(22)
ggplot(total_ctl, aes(tissue, mean_ctl, fill = tissue)) +
  geom_boxplot(outlier.color = NA, show.legend = FALSE, alpha = 0.4) +
  scale_fill_manual(values = c("PBMC" = "#0072B2", "TILN" = "#D55E00")) +
  geom_jitter(color = "grey22", show.legend = FALSE, shape = 21, width = 0.1) +
  stat_compare_means(
    comparisons  = list(c("PBMC", "TILN")),
    method       = "wilcox.test",
    label        = "p.format",
    size         = 3,
    tip.length   = 0,
    bracket.size = 0.8
  ) +
  labs(x = NULL, y = "Mean Cytotoxic UCell Score",
       title = "Total CD4 — Cytotoxic state") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 9 — PBMC vs TILN cell-type ratios

For each CD4 annotation state, we compute a per-patient log2(PBMC / TILN)
frequency ratio as a measure of tissue enrichment. Values above 0 indicate
preferential representation in blood; values below 0 indicate tumour
enrichment. A pseudocount of 1 is added to both numerator and denominator
before log2 transformation. Patient × state combinations with fewer than 5
total cells across both tissues are excluded.

## 9.1 Total CD4 cells

Corresponding to Fig 1F

```{r ratio-total}
rat_int <- compute_pbmc_tiln_ratio(
  metadata       = int@meta.data,
  annotation_col = "annotation",
  patient_col    = "patient",
  tissue_col     = "tissue"
)

cat("Patient × state pairs retained:\n")
print(table(rat_int$annotation, useNA = "ifany"))

plot_pbmc_tiln_ratio(
  ratio_df       = rat_int,
  annotation_col = "annotation",
  palette        = cd4_annotation_palette,
  title          = "PBMC / TILN ratio — total CD4"
)
```

## 9.2 Tumor-specific CD4 cells

Corresponding to Fig 1H

```{r ratio-tumor-specific}
tis_in_proj <- unique(as.character(proj$tissue))

if (length(tis_in_proj) > 1 && all(c("PBMC", "TILN") %in% tis_in_proj)) {
  rat_proj <- compute_pbmc_tiln_ratio(
    metadata       = proj@meta.data,
    annotation_col = "functional.cluster",
    patient_col    = "patient",
    tissue_col     = "tissue"
  )

  cat("Patient × state pairs retained:\n")
  print(table(rat_proj$functional.cluster, useNA = "ifany"))

  plot_pbmc_tiln_ratio(
    ratio_df       = rat_proj,
    annotation_col = "functional.cluster",
    palette        = cd4_annotation_palette,
    title          = "PBMC / TILN ratio — tumor-specific CD4"
  )
} else {
  cat(
    "Tumor-specific CD4 cells only contain tissue(s): ",
    paste(tis_in_proj, collapse = ", "),
    " — PBMC/TILN ratio not computed.\n",
    sep = ""
  )
}
```


# 9.3 — KLRG1+ vs KLRG1- CTL proportion (our data)



```{r klrg1-classification}
DefaultAssay(int) <- "integrated"

require(scGate)
my_scGate_model <- gating_model(name = "KLRG1",
                                signature = "KLRG1") 
int <- scGate(int,
              model = my_scGate_model,
              assay = "integrated",
              ncores = 4)




int$KLRG1_expression <- ifelse(int$is.pure == "Pure",
                               "KLRG1+", "KLRG1-")
DimPlot(int, group.by = "KLRG1_expression")
```
Corresponding to Figure 2A (left)

```{r ratio-klrg1-our-data}

rat_klrg1 <- compute_klrg1_cytotoxic_ratio(
  metadata = int@meta.data,
  patient_col = "patient",
  state_col = "annotation",
  cytotoxic_label = "Cytotoxic",
  klrg1_col = "KLRG1_expression",
  min_cells = 10,
  pseudocount = 1
)

ratio_boxplot_klrg1 <- rat_klrg1 %>%
  ggplot(aes(KLRG1_expression, ratio_log2, fill = KLRG1_expression)) +
  geom_boxplot(
    outlier.color = NA,
    show.legend = FALSE,
    alpha = 0.4
  ) +
  scale_fill_manual(values = c("KLRG1+" = "#0072B2", "KLRG1-" = "#D55E00")) +
  geom_jitter(
    color = "grey22",
    show.legend = FALSE,
    shape = 21,
    width = 0.1,
    alpha = 0.8
  ) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  labs(
    x = "KLRG1 compartment",
    y = "log2 ratio Cytotoxic / Other",
    title = "KLRG1 compartments — total CD4"
  ) +
  ggpubr::theme_classic2() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank()
  )

ratio_boxplot_klrg1
```


# 10 — Cytotoxic CD4 marker discovery

We identify DEGs for
Cytotoxic CD4 cells versus all other annotated CD4 states, then classify
markers as cell-surface receptors using GO molecular-function and
cellular-component annotations.

## 10.1 Differential expression: Cytotoxic vs rest

```{r ctl-markers-find}
DefaultAssay(int) <- "RNA"
Idents(int) <- "annotation"

deg_markers_df <- FindMarkers(
  int,
  ident.1 = "Cytotoxic",
  assay = "RNA",
  test.use = "bimod",
  min.pct = 0.4,
  min.diff.pct = 0.2,
  min.cells.feature = 10,
  verbose = TRUE
)

deg_markers_df <- deg_markers_df %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::filter(!grepl("^HLA|^TR[AB]", gene))

cat("Number of Cytotoxic-vs-rest DEGs after filtering: ", nrow(deg_markers_df), "\n", sep = "")
head(deg_markers_df)
```

## 10.2 Annotate membrane receptor markers

```{r ctl-markers-classify}
source(here::here("scRNA-seq", "markers_classify.R"))

# Extract a clean gene-symbol vector from the DEG table.
gene_symbols <- unique(deg_markers_df$gene)

# Classify symbols by receptor function and membrane localization.
gene_classification <- classify_gene_symbols_surface_receptors(
  gene_symbols = gene_symbols
)

# Add membrane_receptor labels back to the DEG table.
deg_markers_df <- annotate_marker_table_with_receptor_class(
  marker_table = deg_markers_df,
  membrane_receptor_genes = gene_classification$membrane_receptor_genes,
  gene_col = "gene",
  class_col = "membrane_receptor",
  positive_label = "yes",
  negative_label = "no"
)

cat("Receptor class summary:\n")
print(table(deg_markers_df$membrane_receptor, useNA = "ifany"))
```

## 10.3 Plot 20 top Cytotoxic markers

Corresponding to Fig 1J

```{r ctl-markers-plot, fig.width=7, fig.height=5}
bar_cols <- c(
  "Cell-surface receptor" = "#0B7285",
  "Other genes" = "#C5CCD3"
)

top <- 20

gene_markers_ctl <- deg_markers_df %>%
  dplyr::mutate(
    receptor_class = ifelse(membrane_receptor == "yes", "Cell-surface receptor", "Other genes"),
    receptor_class = factor(receptor_class, levels = c("Other genes", "Cell-surface receptor"))
  ) %>%
  dplyr::arrange(desc(avg_log2FC)) %>%
  dplyr::slice_head(n = top) %>%
  dplyr::mutate(gene = forcats::fct_reorder(gene, avg_log2FC)) %>%
  ggplot(aes(x = avg_log2FC, y = gene, fill = receptor_class)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  geom_text(
    aes(label = round(avg_log2FC, 2)),
    hjust = -0.1,
    size = 3,
    color = "grey20"
  ) +
  scale_fill_manual(values = bar_cols) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    fill = "Gene class",
    title = "Top Cytotoxic CD4 marker genes"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9)
  )

gene_markers_ctl
```

```{r ctl-markers-save}
out_ctl <- file.path(cache_dir, "CD4_Cytotoxic_vs_rest_markers.rds")
saveRDS(deg_markers_df, out_ctl)
cat("Saved Cytotoxic marker table: ", out_ctl, "\n", sep = "")
```

---

# 11 — Save reproducible objects

```{r save-objects}
out_int  <- file.path(cache_dir, "cd4_total_integrated_manual.rds")
out_pred <- file.path(cache_dir, "Projectils_prediction_totalCD4.rds")
out_proj <- file.path(cache_dir, "tumor_specific_projection.rds")

saveRDS(int, out_int)
saveRDS(proj, out_proj)

cat("Saved integrated CD4 object: ", out_int, "\n", sep = "")
cat("Saved ProjecTILs predictions: ", out_pred, "\n", sep = "")
cat("Saved tumor-specific projection: ", out_proj, "\n", sep = "")
```

---

# Session information

```{r session-info}
sessionInfo()
```
