---
title: "scRNA-seq Data Processing — Cenerenti et al."
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script reproduces the scRNA-seq/CITE-seq data processing for
Cenerenti et al. starting from the raw feature/barcode matrices deposited
on Zenodo. It covers:

1. Downloading and unzipping the count matrices and sample metadata
2. Quality-control filtering
3. Hashtag-based demultiplexing to assign cells to their sample of origin
4. Merging with the experimental metadata spreadsheet
5. Saving a single merged Seurat object ready for downstream analysis

Raw data: <https://doi.org/10.5281/zenodo.18683366>

The output file `cache/all_samples_merged.rds` is a Seurat object containing
all captures, with per-cell metadata (patient, HLA allele, tissue, antigen
specificity, T cell subset) already attached.

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(Seurat)   # >= 4.0
library(openxlsx) # read/write metadata spreadsheet (.xlsx)
library(ggplot2)
library(patchwork)
library(tidyverse)
```

```{r paths}
# All outputs are written here; the directory already exists in the repo.
cache_dir <- here::here("scRNA-seq", "cache")
```

---

# 1 — Download data from Zenodo

Files are downloaded only when not already present, so re-running this
chunk is safe.

The Zenodo record contains one zip archive per capture (C0.zip – C8.zip)
and a `metadata.xlsx` file that maps every hashtag barcode to its sample.

```{r download}
zenodo_url <- "https://zenodo.org/records/18683366/files"

# Files expected in the record
zenodo_files <- c(paste0("C", 0:8, "_sample_feature_bc_matrix.zip"), "metadata.xlsx")

for (f in zenodo_files) {
  dest <- file.path(cache_dir, f)
  if (!file.exists(dest)) {
    cat("Downloading ", f, " ...\n", sep = "")
    download.file(
      url      = paste0(zenodo_url, "/", f, "?download=1"),
      destfile = dest,
      mode     = "wb",
      quiet    = FALSE
    )
  } else {
    cat(f, " already present — skipping download.\n", sep = "")
  }
}
```

---

# 2 — Load metadata spreadsheet

`metadata.xlsx` has one row per hashtag and provides all the biological
context needed to annotate cells after demultiplexing.

```{r load-metadata}
meta <- openxlsx::read.xlsx(file.path(cache_dir, "metadata.xlsx"))

# Preview
print(meta)
```

```{r parse-metadata}
# Derive a clean numeric hashtag index ("Hashtag 1" -> 1, etc.)
# and biological annotation fields used later.
meta <- meta %>%
  mutate(
    ht_index = as.integer(str_extract(`N°`, "[0-9]+")),

    # Short antigen label (NA for bulk CD4/CD8 populations)
    antigen = case_when(
      str_detect(Population, regex("NY-ESO|NYESO", ignore_case = TRUE)) ~ "NYESO",
      str_detect(Population, regex("\\bHA\\b",     ignore_case = TRUE)) ~ "HA",
      str_detect(Population, regex("MAGE",         ignore_case = TRUE)) ~ "MAGE",
      str_detect(Population, regex("\\bTT\\b",     ignore_case = TRUE)) ~ "TT",
      TRUE ~ NA_character_
    ),

    # Antigen-specificity class
    # NY-ESO-1 and MAGE-A3 are tumour antigens;
    # HA (influenza) and TT (tetanus toxin) are pathogen-derived controls.
    specificity = case_when(
      antigen %in% c("NYESO", "MAGE") ~ "tumor_specific",
      antigen %in% c("HA",    "TT")   ~ "virus_specific",
      TRUE                            ~ "non_specific"
    ),

    # T cell subset — all tetramer-enriched populations are CD4+
    T_subset = if_else(Population == "CD8", "CD8", "CD4"),

    # Human-readable HTO label that will be set as rowname of the HTG assay
    # and later appear in the MULTI_ID metadata column.
    # Format: <Capture>-HT<N>-<CD4|CD8>[-<antigen>]-<Type>
    ht_label = if_else(
      is.na(antigen),
      paste(Capture, paste0("HT", ht_index), T_subset, Type, sep = "-"),
      paste(Capture, paste0("HT", ht_index), T_subset, antigen, Type, sep = "-")
    )
  )

# Show the derived labels for verification
meta %>%
  select(Capture, `N°`, Patient, HLA, Population, Type,
         antigen, specificity, T_subset, ht_label)
```

---

# 3 — Load feature/barcode matrices (unzip on the fly)

Each capture zip is extracted into `cache/` before reading with
`Seurat::Read10X`. The HTO assay row names are set to the human-readable
labels derived from `metadata.xlsx`.

```{r load-matrices}
captures <- sort(unique(meta$Capture))   # "C0", "C1", ..., "C8"

samples <- lapply(setNames(captures, captures), function(cap) {

  # ── 1. Unzip on the fly ────────────────────────────────────────────────────
  # Zips are named <cap>_sample_feature_bc_matrix.zip and extract directly to
  # a folder of the same name (no extra subdirectory).
  zip_path <- file.path(cache_dir, paste0(cap, "_sample_feature_bc_matrix.zip"))
  mat_dir  <- file.path(cache_dir, paste0(cap, "_sample_feature_bc_matrix"))

  if (!dir.exists(mat_dir)) {
    cat("Unzipping ", basename(zip_path), " ...\n", sep = "")
    unzip(zip_path, exdir = cache_dir)
  }

  # ── 2. Read gene-expression + hashtag matrices ─────────────────────────────
  mat <- Seurat::Read10X(mat_dir)

  seu <- CreateSeuratObject(
    counts  = mat$`Gene Expression`,
    project = cap
  )

  # ── 3. Attach hashtag (HTG) assay with labelled row names ─────────────────
  ht_meta <- meta %>%
    filter(Capture == cap) %>%
    arrange(ht_index)

  # The number of HTO features in the matrix must match the metadata rows
  stopifnot(nrow(mat$`Antibody Capture`) == nrow(ht_meta))

  rownames(mat$`Antibody Capture`) <- ht_meta$ht_label
  seu[["HTG"]] <- CreateAssayObject(counts = mat$`Antibody Capture`)

  cat(cap, ": ", ncol(seu), " cells loaded\n", sep = "")

  # ── 4. Remove extracted folder to save disk space ─────────────────────────
  unlink(mat_dir, recursive = TRUE)
  cat(cap, ": extracted folder removed.\n", sep = "")

  return(seu)
})
```

---

# 4 — Quality control

## 4.1 Compute QC metrics

```{r qc-metrics}
for (i in seq_along(samples)) {
  samples[[i]][["percent.mt"]] <- PercentageFeatureSet(
    samples[[i]], pattern = "^MT-"
  )
}
```

## 4.2 Visualise before filtering

```{r qc-violin, fig.height=4.5, fig.width=10}
# VlnPlot with multiple features already returns a patchwork object
for (s in names(samples)) {
  p <- VlnPlot(samples[[s]],
               features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
               ncol = 3, pt.size = 0) &
         theme(legend.position = "none")
  print(p + patchwork::plot_annotation(title = s))
}
```

## 4.3 Filter low-quality cells

Thresholds match `Integrated_analysis.Rmd` and `Added_samples_analysis.Rmd`:
- `nFeature_RNA > 300` — removes empty droplets / very low-complexity cells
- `percent.mt < 8`     — removes broken or dying cells

```{r qc-filter}
for (i in seq_along(samples)) {
  n_before <- ncol(samples[[i]])
  samples[[i]] <- subset(
    samples[[i]],
    subset = nFeature_RNA > 300 & percent.mt < 8
  )
  cat(names(samples)[i], ": ", n_before,
      " -> ", ncol(samples[[i]]), " cells after QC filter\n", sep = "")
}
```

---

# 5 — Hashtag demultiplexing

Cells are assigned to their sample of origin based on HTO signal.
`MULTIseqDemux` on CLR-normalised counts was used in the original
analysis and is reproduced here with the same parameters.

## 5.1 Normalise HTG assay and call sample identity

```{r demux-call}
for (i in seq_along(samples)) {

  # CLR normalisation of the hashtag assay
  samples[[i]] <- NormalizeData(
    samples[[i]],
    assay                = "HTG",
    normalization.method = "CLR"
  )

  # Sample assignment per cell
  samples[[i]] <- MULTIseqDemux(
    samples[[i]],
    assay      = "HTG",
    autoThresh = TRUE,
    maxiter    = 20
  )

  # Re-level: HTO labels first (ordered), then Negative / Doublet
  samples[[i]]$MULTI_ID <- factor(
    samples[[i]]$MULTI_ID,
    levels = c(rownames(samples[[i]][["HTG"]]), "Negative", "Doublet")
  )

  cat(names(samples)[i], " - demultiplexing:\n", sep = "")
  print(table(samples[[i]]$MULTI_ID))
}
```

## 5.2 Inspect HTO signal distributions

```{r demux-ridge, fig.height=10, fig.width=15}
for (i in seq_along(samples)) {
  plot(
    RidgePlot(
      samples[[i]],
      assay    = "HTG",
      features = rownames(samples[[i]][["HTG"]]),
      ncol     = 3
    ) + ggtitle(names(samples)[i])
  )
}
```

## 5.3 Remove doublets and negatives

```{r demux-filter}
for (i in seq_along(samples)) {
  n_before <- ncol(samples[[i]])
  samples[[i]] <- subset(
    samples[[i]],
    subset = MULTI_ID %in% c("Doublet", "Negative"),
    invert = TRUE
  )
  cat(names(samples)[i], ": ", n_before,
      " -> ", ncol(samples[[i]]),
      " cells after doublet/negative removal\n", sep = "")
}
```

---

# 6 — Attach experimental metadata

Each cell is annotated with patient, HLA allele, antigen, antigen-specificity
class, T cell subset (CD4/CD8) and tissue of origin (PBMC / TILN). Values
are looked up from `metadata.xlsx` via the MULTI_ID hashtag label — no
hardcoding per capture.

```{r add-metadata}
for (cap in names(samples)) {

  # Build a lookup keyed on the HTO label
  lut <- meta %>%
    filter(Capture == cap) %>%
    select(ht_label, Patient, HLA, antigen, specificity, T_subset, Type) %>%
    column_to_rownames("ht_label")

  idx <- match(as.character(samples[[cap]]$MULTI_ID), rownames(lut))

  samples[[cap]]$patient     <- lut$Patient[idx]
  samples[[cap]]$HLA         <- lut$HLA[idx]
  samples[[cap]]$antigen     <- lut$antigen[idx]
  samples[[cap]]$specificity <- lut$specificity[idx]
  samples[[cap]]$T_subset    <- lut$T_subset[idx]
  samples[[cap]]$tissue      <- lut$Type[idx]   # PBMC or TILN
}

# Spot-check: one row per hashtag for a representative capture
samples[["C0"]]@meta.data %>%
  distinct(MULTI_ID, patient, HLA, antigen, specificity, T_subset, tissue) %>%
  arrange(MULTI_ID)
```

---

# 7 — Merge all captures and normalise

All nine captures are merged into a single Seurat object. Log-normalisation
(scale factor 10 000) is applied so the object is ready for variable-feature
selection and integration in downstream scripts.

```{r merge}
merged <- merge(
  samples[[1]],
  y          = samples[-1],
  merge.data = TRUE
)

merged <- NormalizeData(merged)

# Summary
cat("Merged object: ", ncol(merged), " cells x ", nrow(merged), " genes\n", sep = "")
cat("Captures : ", paste(sort(unique(merged$orig.ident)), collapse = ", "), "\n", sep = "")
cat("Patients : ", paste(sort(na.omit(unique(merged$patient))), collapse = ", "), "\n", sep = "")
cat("T subsets: ", paste(table(merged$T_subset), collapse = " | "), "\n", sep = "")
cat("Tissues  : ", paste(table(merged$tissue),   collapse = " | "), "\n", sep = "")
```

---

# 8 — Save

```{r save}
out_path <- file.path(cache_dir, "all_samples_merged.rds")
saveRDS(merged, out_path)
cat("Saved: ", out_path, "\n", sep = "")
```

---

# Session information

```{r session-info}
sessionInfo()
```
