---
title: "Zheng et al. Multi-cancer CD4 Annotation with ProjecTILs"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script reproduces the annotation component of the Zheng et al. multi-cancer
cohort workflow, restricted to:

1. Loading the processed Zheng object
2. Filtering CD4 T cells
3. Annotating CD4 cells with ProjecTILs
4. Saving an annotated Zheng Seurat object as `.rds`

## Input

- `scRNA-seq/cache/Zheng_2021_34914499_whole.rds`
- Generated by `scRNA-seq/2.0.Zheng_processing.Rmd`

## Outputs

- `scRNA-seq/cache/Zheng_2021_34914499_CD4_annotated.rds`
- `scRNA-seq/cache/Zheng_2021_34914499_CD4_projectils_metadata.rds`

---

# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(Seurat)
library(ProjecTILs)
library(tidyverse)
library(data.table)
library(here)
library(ggpubr)
```

```{r paths}
cache_dir <- here::here("scRNA-seq", "cache")
in_rds    <- file.path(cache_dir, "Zheng_2021_34914499_whole.rds")
out_rds   <- file.path(cache_dir, "Zheng_2021_34914499_CD4_annotated.rds")
out_meta  <- file.path(cache_dir, "Zheng_2021_34914499_CD4_projectils_metadata.rds")

if (!file.exists(in_rds)) {
  stop("Input not found: ", in_rds)
}
```

---

# 1 - Load Zheng object and define CD4 cells

```{r load-zheng}
zheng <- readRDS(in_rds)

if (!inherits(zheng, "Seurat")) {
  stop("Input file does not contain a Seurat object: ", in_rds)
}

cat("Loaded Zheng object: ", ncol(zheng), " cells x ", nrow(zheng), " genes\n", sep = "")
```

```{r derive-cd4-flag}
# if authors_Tcell is absent, infer it from meta.cluster labels.
if (!"authors_Tcell" %in% colnames(zheng@meta.data)) {
  if (!"meta.cluster" %in% colnames(zheng@meta.data)) {
    stop("Cannot derive CD4 cells: neither 'authors_Tcell' nor 'meta.cluster' found in metadata.")
  }
  zheng$authors_Tcell <- ifelse(grepl("^CD4", zheng$meta.cluster), "CD4T", "CD8T")
}

if (!"Sample" %in% colnames(zheng@meta.data)) {
  if ("orig.ident" %in% colnames(zheng@meta.data)) {
    zheng$Sample <- as.character(zheng$orig.ident)
  } else {
    stop("Metadata column 'Sample' is missing and could not be inferred.")
  }
}

# Retain onlyh CD4+ T cells
cd4 <- subset(zheng, subset = authors_Tcell == "CD4T")
cat("CD4 cells retained: ", ncol(cd4), "\n", sep = "")
print(table(cd4$Sample))
rm(zheng)
gc()
```

---

# 2 - Load ProjecTILs CD4 reference map

```{r load-reference}
ref_url  <- "https://figshare.com/ndownloader/files/43794750"
ref_path <- file.path(cache_dir, "CD4T_human_ref_v2.rds")

if (!file.exists(ref_path)) {
  cat("Downloading ProjecTILs CD4 reference map ...\n")
  download.file(ref_url, destfile = ref_path, mode = "wb", quiet = FALSE)
}

cd4ref <- ProjecTILs::load.reference.map(ref_path)
UMAPPlot(cd4ref) + ggtitle("ProjecTILs CD4 reference map")
```

---

# 3 - Annotate CD4 cells with ProjecTILs

Annotation is run per sample.
Samples with fewer than 10 CD4 cells are skipped.

```{r projectils-annotate}
samples <- unique(as.character(cd4$Sample))
pred_list <- list()

for (i in seq_along(samples)) {
  s <- samples[i]
  obj <- subset(cd4, subset = Sample == s)

  cat("[", i, "/", length(samples), "] ", s, " -> ", ncol(obj), " CD4 cells\n", sep = "")

  if (ncol(obj) < 10) {
    cat("  Skipping (fewer than 10 CD4 cells).\n")
    next
  }

  pred <- ProjecTILs.classifier(
    query = obj,
    ref = cd4ref,
    filter.cells = FALSE,
    ncores = 1
  )

  pred_md <- pred@meta.data %>%
    tibble::rownames_to_column("cellid") %>%
    mutate(Sample = s)

  pred_list[[s]] <- pred_md

  rm(obj, pred, pred_md)
  gc()
}

if (length(pred_list) == 0) {
  stop("No samples were annotated. Check CD4 filtering and sample sizes.")
}

pred_md_all <- data.table::rbindlist(pred_list, fill = TRUE)
cat("Annotated CD4 cells: ", nrow(pred_md_all), "\n", sep = "")
```

---

# 4 - Merge ProjecTILs labels back to full Zheng object

```{r merge-annotations}
md <- cd4@meta.data %>%
  tibble::rownames_to_column("cellid")

# Keep only ProjecTILs-derived columns and prefix them to avoid name conflicts.
proj_cols <- setdiff(colnames(pred_md_all), c("cellid", "Sample"))
pred_for_join <- pred_md_all %>%
  dplyr::select(cellid, "functional.cluster")

md2 <- md %>%
  left_join(pred_for_join, by = "cellid") %>%
  tibble::column_to_rownames("cellid")

cd4@meta.data <- md2

cat("Cells with ProjecTILs CD4 annotation:\n")
if ("functional.cluster" %in% colnames(cd4@meta.data)) {
  print(table(cd4$functional.cluster, useNA = "ifany"))
} else {
  cat("Column functional.cluster not found in predictions.\n")
}
```

---

# 5 - Save outputs

```{r save}
saveRDS(cd4, out_rds)
saveRDS(pred_md_all, out_meta)

cat("Saved annotated Zheng object: ", out_rds, "\n", sep = "")
cat("Saved ProjecTILs metadata: ", out_meta, "\n", sep = "")
```

---

# 6 - Zheng compositional

Corresponding to Fig. 1I

```{r zheng-compositional}
if (!"functional.cluster" %in% colnames(cd4@meta.data)) {
  stop("Column functional.cluster is required for compositional analysis.")
}

md_zheng <- cd4@meta.data

if (!"Tissue" %in% colnames(md_zheng)) {
  stop("Column Tissue is required for compositional analysis.")
}

sel_ct <- c("CD4.CTL_GNLY", "CD4.Tfh")
pseudocount <- 0.5
min_cells_selct <- 5

md_zheng <- md_zheng %>%
  filter(Tissue %in% c("P", "T", "N")) %>%
  mutate(
    Tissue = factor(
      Tissue,
      levels = c("T", "N", "P"),
      labels = c("Tumor", "Normal", "PBMC")
    )
  )

# CLR across sel_ct within each sample (patient x tissue)
rat <- md_zheng %>%
  filter(functional.cluster %in% sel_ct) %>%
  count(patient, functional.cluster, Tissue, name = "ncells") %>%
  complete(
    patient,
    functional.cluster = factor(sel_ct, levels = sel_ct),
    Tissue = c("Tumor", "Normal", "PBMC"),
    fill = list(ncells = 0)
  ) %>%
  group_by(patient, Tissue) %>%
  mutate(total_selct = sum(ncells)) %>%
  filter(total_selct >= min_cells_selct) %>%
  mutate(
    prop = (ncells + pseudocount) / sum(ncells + pseudocount),
    clr = log(prop) - mean(log(prop)),
  ) %>%
  ungroup() %>% 
  mutate(Tissue = factor(Tissue,
                         levels = c("Tumor", "Normal", "PBMC")))

comps <- list(c("Tumor", "Normal"),
              c("Tumor", "PBMC"),
              c("Normal", "PBMC"))

set.seed(22)
ratio_boxplot_zheng <- rat %>% 
  ggplot(aes(Tissue, clr, fill = Tissue)) +
  geom_boxplot(outlier.color = NA,
               show.legend = F,
               alpha = 0.4) +
  scale_fill_manual(values = c("Tumor" = "#C23B22",
                               "Normal" = "#2F5D8A",
                               "PBMC" = "#1A9A8A")) +
  geom_jitter(color = "grey22",
              show.legend = FALSE,
              shape = 21,
              width = 0.1,
              alpha = 0.8) +
  ggpubr::stat_compare_means(
    comparisons = comps,
    method = "wilcox.test",
    paired = FALSE,
    label = "p.adj.format",
    p.adjust.method = "BH",
    size = 3.5,
    tip.length = 0,
    bracket.size = 0.3
  ) +
  geom_hline(yintercept = 0,
             color = "grey40",
             linetype = "dashed") +
  labs(x = "Tissue",
       y = "Centered-log ratio abundance",
       fill = "Tissue") +
  facet_wrap(~functional.cluster) +
  ggpubr::theme_classic2() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1))
ratio_boxplot_zheng

```



# 7 - Zheng markers

```{r zheng-markers-setup}
if (!"functional.cluster" %in% colnames(cd4@meta.data)) {
  stop("Column functional.cluster is required for marker analysis.")
}

Idents(cd4) <- "functional.cluster"
```

## 7.1 Marker heatmap

Corresponding to Suppl Figure 1G

```{r zheng-marker-heatmap}
cd4_cytotoxic_markers <- c("GZMB", "GNLY", "NKG7", "FGFBP2", "PRF1", "CST7", "GZMA")
tfh_markers <- c("CXCR5", "PDCD1", "CXCL13", "IL21", "TOX")

feats_heatmap <- c(cd4_cytotoxic_markers, tfh_markers)
sel_ct <- c("CD4.CTL_GNLY", "CD4.Tfh")

cd4_sub <- cd4[,cd4$functional.cluster %in%
                         sel_ct &
                         cd4$Tissue != "LN"]
zheng_heatmap_markers <- ProjecTILs::celltype.heatmap(
  cd4_sub,
  assay = "RNA",
  slot = "data",
  genes = feats_heatmap,
  cluster.col = "functional.cluster",
  cluster_genes = FALSE,
  min.cells = 5,
  group.by = c("Tissue", "cancerType")
)

```

## 7.2 Differential markers for CD4.CTL_GNLY

```{r zheng-deg-markers}
if (!"CD4.CTL_GNLY" %in% unique(Idents(cd4))) {
  stop("CD4.CTL_GNLY cluster not found in functional.cluster annotations.")
}

deg_markers_df <- FindMarkers(
  object = cd4,
  ident.1 = "CD4.CTL_GNLY",
  assay = "RNA",
  test.use = "bimod",
  min.pct = 0.4,
  min.diff.pct = 0.2,
  min.cells.feature = 10,
  verbose = TRUE
) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::filter(!grepl("^HLA|^TR[AB]", gene))

source(here::here("scRNA-seq", "markers_classify.R"))

gene_symbols <- unique(deg_markers_df$gene)
gene_classification <- classify_gene_symbols_surface_receptors(gene_symbols = gene_symbols)

deg_markers_df <- annotate_marker_table_with_receptor_class(
  marker_table = deg_markers_df,
  membrane_receptor_genes = gene_classification$membrane_receptor_genes,
  gene_col = "gene",
  class_col = "membrane_receptor",
  positive_label = "yes",
  negative_label = "no"
)

table(deg_markers_df$membrane_receptor)
```

## 7.3 Top 20 CTL marker barplot

Corresponding to Suppl Fig 1H

```{r zheng-marker-barplot}
bar_cols <- c(
  "Cell-surface receptor" = "#0B7285",
  "Other genes" = "#C5CCD3"
)

gene_markers_ctl_zheng <- deg_markers_df %>%
  mutate(
    receptor_class = ifelse(membrane_receptor == "yes", "Cell-surface receptor", "Other genes"),
    receptor_class = factor(receptor_class, levels = c("Other genes", "Cell-surface receptor"))
  ) %>%
  arrange(desc(avg_log2FC)) %>%
  slice_head(n = 20) %>%
  mutate(gene = forcats::fct_reorder(gene, avg_log2FC)) %>%
  ggplot(aes(x = avg_log2FC, y = gene, fill = receptor_class)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  geom_text(aes(label = round(avg_log2FC, 2)), hjust = -0.1, size = 3, color = "grey20") +
  scale_fill_manual(values = bar_cols) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  labs(
    x = "log2 Fold Change",
    y = "Gene",
    fill = "Gene class",
    title = "Top Cytotoxic CD4+ Marker Genes (Zheng)"
  ) +
  ggpubr::theme_classic2() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9)
  )

gene_markers_ctl_zheng
```

---

# 8 - Proportions of CTL/non-CTL within KLRG1+ and KLRG1- (Zheng)

```{r}
source(here::here("scRNA-seq", "klrg1_ratio_utils.R"))
```

Consider splitting by sample of origin for better smoothing
```{r klrg1-classification}
require(scGate)
my_scGate_model <- gating_model(name = "KLRG1",
                                signature = "KLRG1") 
cd4 <- scGate(cd4,
              model = my_scGate_model,
              ncores = 4)

cd4$KLRG1_expression <- ifelse(cd4$is.pure == "Pure",
                               "KLRG1+", "KLRG1-")
```

Corresponding to Figure 2A (right)

```{r ratio-klrg1-zheng}

patient_col <- dplyr::case_when(
  "patient" %in% colnames(cd4@meta.data) ~ "patient",
  "Individual" %in% colnames(cd4@meta.data) ~ "Individual",
  TRUE ~ "Sample"
)

rat_klrg1_zheng <- compute_klrg1_cytotoxic_ratio(
  metadata = cd4@meta.data,
  patient_col = patient_col,
  state_col = "functional.cluster",
  cytotoxic_label = "CD4.CTL_GNLY",
  klrg1_col = "KLRG1_expression",
  min_cells = 10,
  pseudocount = 1
)

ratio_boxplot_klrg1_zheng <- rat_klrg1_zheng %>%
  ggplot(aes(KLRG1_expression, ratio_log2, fill = KLRG1_expression)) +
  geom_boxplot(
    outlier.color = NA,
    show.legend = FALSE,
    alpha = 0.4
  ) +
  scale_fill_manual(values = c("KLRG1+" = "#0072B2", "KLRG1-" = "#D55E00")) +
  geom_jitter(
    color = "grey22",
    show.legend = FALSE,
    shape = 21,
    width = 0.1,
    alpha = 0.8
  ) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  labs(
    x = "KLRG1 compartment",
    y = "log2 ratio Cytotoxic / Other",
    title = "KLRG1 compartments — Zheng CD4"
  ) +
  ggpubr::theme_classic2() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank()
  )

ratio_boxplot_klrg1_zheng
```

---

# Session info

```{r session-info}
sessionInfo()
```
