---
title: "Zheng et al. Multi-cancer CD4 Data Processing"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script downloads, loads, and processes the Zheng et al. 2021 pan-cancer T cell dataset
for integration with the KLRG1 CD4-CTL reproducibility project. The script:

1. Downloads raw data from Zenodo if needed
2. Loads SCE objects and converts to Seurat format
3. Performs QC filtering (ribosomal content, mitochondrial content, feature/count metrics)
4. Standardizes gene symbols
5. Saves the merged, high-quality object

## Dataset Information

- **Citation**: Zheng et al. (2021) - Pan-cancer single-cell landscape of tumor-infiltrating T cells
- **Zenodo DOI**: https://zenodo.org/record/5461803
- **Output**: `scRNA-seq/cache/Zheng_2021_34914499_whole.rds` (CD4+CD8 combined, high-quality cells)

---

# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(SingleCellExperiment)
library(STACAS)
library(SignatuR)
library(openxlsx)


read_excel_allsheets <- function(filename) {
  sheets <- openxlsx::getSheetNames(filename)
  x <- lapply(sheets, function(sheet_name) {
    openxlsx::read.xlsx(filename, sheet = sheet_name)
  })
  names(x) <- sheets
  x
}


options(timeout = 4000) # increase timeout for downloading
set.seed(321)
```

## Define parameters

```{r define_parameters}
# Dataset identifiers
ds_name <- "Zheng_2021_34914499"
dataset_id <- "ZheL"
species <- "human"

# QC filtering parameters
min_cells_per_sample <- 200
```

## Set up data paths

```{r setup_paths}
# Reference gene annotations (for gene symbol standardization) from STACAS
data("EnsemblGeneTable.Hs")

# Data paths
data_dir <- here("scRNA-seq", "cache")
input_data_path <- file.path(data_dir, "Zheng_raw")

# Create directories if needed
dir.create(input_data_path, showWarnings = FALSE, recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
```

# 1 - Download data

## Zenodo download (skip if data already present)

The data is downloaded from Zenodo (DOI: https://zenodo.org/record/5461803). 
If already downloaded, this chunk is skipped.

```{r download_zenodo, eval=FALSE}
library(zen4R)

# Create subdirectory for Zenodo data
zenodo_dir <- file.path(input_data_path, "Zenodo")
dir.create(zenodo_dir, showWarnings = FALSE, recursive = TRUE)

# Download if not already present
if (length(list.files(zenodo_dir)) == 0) {
  cat("Downloading Zheng dataset from Zenodo...\n")
  zenodo_doi <- "10.5281/zenodo.5461803"
  
  zen4R::download_zenodo(doi = zenodo_doi, path = zenodo_dir)
  
  # Untar expression datasets and metadata
  tar_zenodo_files <- list.files(zenodo_dir, "tar", full.names = TRUE) %>% 
    grep("expression|meta", ., value = TRUE, ignore.case = TRUE)
  
  lapply(tar_zenodo_files, untar)
  cat("Download complete.\n")
} else {
  cat("Data already present in", zenodo_dir, "\n")
}
```



# 2 - Load data

## Load SCE objects and convert to Seurat

Zheng et al. provides SingleCellExperiment (SCE) objects with both CD4 and CD8 T cells 
split by dataset. We load all available datasets and convert them to Seurat format,
keeping only those with raw count data.

```{r load_sce_objects}
# Initialize lists to store file paths and metadata
file_paths <- list()
file_names <- list()
dataset_names <- list()

# Collect paths for CD4 and CD8 data
for (cd_type in c("CD4", "CD8")) {
  data_subdir <- file.path(input_data_path, "Zenodo", "expression", cd_type, "byDataset")
  
  if (!dir.exists(data_subdir)) {
    warning("Directory not found: ", data_subdir)
    next
  }
  
  # Get all .rds files
  file_paths[[cd_type]] <- list.files(
    path = data_subdir,
    pattern = "sce.rds$",
    full.names = TRUE
  )
  
  # Get file names (without .rds extension)
  file_names[[cd_type]] <- list.files(
    path = data_subdir,
    pattern = "sce.rds$"
  )
  
  # Dataset names (remove .sce.rds suffix)
  dataset_names[[cd_type]] <- gsub(".sce.rds", "", file_names[[cd_type]])
  
  # Assign names to the file path vector
  names(file_paths[[cd_type]]) <- dataset_names[[cd_type]]
}

cat("Found", length(file_paths$CD4), "CD4 datasets and", 
    length(file_paths$CD8), "CD8 datasets\n")

# Verify that CD4 and CD8 have matching dataset names
if (!identical(dataset_names$CD4, dataset_names$CD8)) {
  warning("Some datasets do not have both CD4 and CD8 samples")
}

# Load SCE objects and convert to Seurat
# Keep only datasets with raw count data
seurat_list <- list()

for (dataset_name in dataset_names$CD4) {
  seurat_list[[dataset_name]] <- list()
  
  for (cd_type in names(dataset_names)) {
    sce <- readRDS(file_paths[[cd_type]][[dataset_name]])
    
    # Check for raw counts
    if ("counts" %in% names(sce@assays)) {
      # Convert SCE to Seurat (without normalized data slot)
      seu <- as.Seurat(sce, data = NULL)
      
      # Rename assay from 'originalexp' to 'RNA' for consistency
      seu <- RenameAssays(seu, originalexp = "RNA")

      # Standardize gene symbols immediately after conversion to Seurat.
      seu <- STACAS::StandardizeGeneSymbols(seu, EnsemblGeneTable.Hs)

      # Rebuild the Seurat object so counts and metadata remain aligned
      # after gene symbol standardization.
      seu <- CreateSeuratObject(
        counts = GetAssayData(seu, assay = "RNA", layer = "counts"),
        meta.data = seu@meta.data,
        assay = "RNA",
        min.cells = 0,
        min.features = 0
      )
      
      # Add CD type to metadata
      seu$cd_type <- cd_type
      
      seurat_list[[dataset_name]][[cd_type]] <- seu
    } else {
      cat(dataset_name, "for", cd_type, "does not have raw counts - skipping\n")
    }
  }
}

# Remove empty list elements
keep_datasets <- names(seurat_list)[sapply(seurat_list, function(x) length(x) > 0)]
seurat_list <- seurat_list[keep_datasets]

cat("Loaded", length(seurat_list), "datasets with raw counts\n")
```

# 3 - Adapt and add metadata

```{r adapt_meta}
cohort_info_file <- file.path(input_data_path, "Zenodo", "cohort_patient_info.xlsx")
clin_info_file <- file.path(
  input_data_path,
  "Zenodo",
  "metaInfo",
  "TableS1.clinicalInfo.P316.txt"
)

if (!file.exists(cohort_info_file)) {
  stop("Missing cohort metadata file: ", cohort_info_file)
}

if (!file.exists(clin_info_file)) {
  stop("Missing clinical metadata file: ", clin_info_file)
}

donor_info <- read_excel_allsheets(cohort_info_file) %>%
  lapply(function(x) {
    names(x) <- tolower(x[1, ])
    x[-1, ]
  })

if ("Cohort Info" %in% names(donor_info)) {
  names(donor_info$`Cohort Info`)[1] <- "cancerType"
}

clin_info <- read.table(
  clin_info_file,
  sep = "\t",
  header = TRUE,
  check.names = FALSE
) %>%
  distinct() %>%
  separate(Patient.ID, into = c("dataset", "patient"), sep = "Study.", remove = FALSE)

extract_barcode <- function(cell_name) {
  pieces <- strsplit(as.character(cell_name), "_")[[1]]
  if (length(pieces) >= 2) {
    pieces[[2]]
  } else {
    sub("[.].*$", "", as.character(cell_name))
  }
}

for (dataset_name in names(seurat_list)) {
  for (cd_type in names(seurat_list[[dataset_name]])) {
    tech <- NA_character_
    if ("Cohort Info" %in% names(donor_info)) {
      tech_idx <- donor_info$`Cohort Info`$dataset == dataset_name
      if (any(tech_idx, na.rm = TRUE) && "technology" %in% names(donor_info$`Cohort Info`)) {
        tech <- donor_info$`Cohort Info`[tech_idx, "technology"][[1]]
      }
    }

    seu <- seurat_list[[dataset_name]][[cd_type]]
    meta <- seu@meta.data

    if (!"Original_cell_names" %in% names(meta)) {
      meta$Original_cell_names <- colnames(seu)
    }

    if (!"patient" %in% names(meta)) {
      meta$patient <- NA_character_
    }

    if (!"orig.ident" %in% names(meta)) {
      meta$orig.ident <- dataset_name
    }

    if (!"batchV" %in% names(meta)) {
      meta$batchV <- NA_character_
    }

    if (!"cellID.uniq" %in% names(meta)) {
      meta$cellID.uniq <- colnames(seu)
    }

    meta <- meta %>%
      mutate(
        Original_cell_names = as.character(Original_cell_names),
        orig.ident = as.character(orig.ident),
        patient = as.character(patient),
        batchV = as.character(batchV),
        Barcode = vapply(Original_cell_names, extract_barcode, character(1)),
        Study = ds_name,
        dataset = dataset_name,
        cd_type = cd_type,
        Individual = patient,
        Patient.ID = paste(dataset, patient, sep = "."),
        Sample = paste(orig.ident, Patient.ID, batchV, sep = "_"),
        cell_ids = paste(dataset_id, cellID.uniq, sep = "."),
        technology = tech
      ) %>%
      rowwise() %>%
      mutate(
        time = ifelse(
          grepl("-D", orig.ident),
          strsplit(orig.ident, "-")[[1]][2],
          NA_character_
        ),
        Treatment = ifelse(
          grepl("pre|post", orig.ident, ignore.case = TRUE),
          tail(strsplit(orig.ident, "[.]")[[1]], 1),
          NA_character_
        )
      ) %>%
      ungroup()

    if (!grepl("this", dataset_name, ignore.case = TRUE)) {
      meta <- meta %>%
        mutate(
          Patient.ID = gsub("LUNG[.]", "NSCLC.", Patient.ID),
          Patient.ID = gsub("BC[.]", "BRCA.", Patient.ID),
          Patient.ID = gsub("Peter2018", "PeterSavas", Patient.ID),
          Patient.ID = gsub("ZiyiLi10X", "LeiZhang2020", Patient.ID),
          Patient.ID = gsub("YaoHe10X", "QimingZhang2019.10X", Patient.ID),
          Patient.ID = gsub("Melanoma", "MELA", Patient.ID),
          Patient.ID = gsub("XiliangWang2019", "YangLiu2020", Patient.ID),
          Patient.ID = gsub("Elham2018", "ElhamAzizi2018", Patient.ID),
          Patient.ID = gsub("Indrop", "InDrop", Patient.ID),
          Patient.ID = gsub("zhangLabSS2[.]CRC", "LeiZhang2018.CRC", Patient.ID),
          Patient.ID = gsub("LeiZhang2020", "LeiZhang2020.10X", Patient.ID),
          Patient.ID = gsub("zhangLabSS2[.]HCC", "ChunhongZheng2017.HCC", Patient.ID),
          Patient.ID = gsub("YaoHeSS2", "QimingZhang2019.SS2", Patient.ID),
          Patient.ID = gsub("129p[a|b]", "129", Patient.ID),
          Patient.ID = gsub("zhangLabSS2[.]NSCLC", "XinyiGuo2018.LUNG", Patient.ID)
        ) %>%
        left_join(clin_info[, -c(2:7)], by = "Patient.ID") %>%
        distinct() %>%
        tibble::column_to_rownames(var = "cell_ids")
    } else {
      meta <- meta %>%
        left_join(clin_info[, -c(1, 2, 4:7)], by = "patient") %>%
        distinct() %>%
        tibble::column_to_rownames(var = "cell_ids")
    }

    seu <- RenameCells(seu, new.names = rownames(meta))
    seu@meta.data <- meta
    seurat_list[[dataset_name]][[cd_type]] <- seu
  }
}
```

# 4 - Merge datasets

```{r merge_objects}
# Flatten the nested dataset/CD4-CD8 structure before merging.
seurat_objects <- unlist(seurat_list, recursive = FALSE)

if (length(seurat_objects) == 0) {
  stop("No Seurat objects were loaded from the Zheng dataset.")
}

merged_seu <- Reduce(
  f = function(x, y) merge(x, y),
  x = seurat_objects
)

individual_metadata <- merged_seu@meta.data %>%
  tibble::rownames_to_column(var = "cell_id") %>%
  distinct(
    Patient.ID,
    cancerType,
    Age,
    Gender,
    pTNM,
    Clinical.Stage,
    BMI,
    Lymphocyte.count..10e9.L.
  )

merged_seu@misc$batch.metadata <- list()
merged_seu@misc$batch.metadata$Individual <- individual_metadata

cat("Merged object has", ncol(merged_seu), "cells and", 
    nrow(merged_seu), "genes\n")
```





# 5 - Quality Control and Filtering

## Calculate QC metrics

```{r calculate_qc_metrics}
library(SignatuR)

# Get mitochondrial and ribosomal gene signatures
mito_genes <- GetSignature(SignatuR$Hs$Compartments$Mito)
ribo_genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)

# Add QC metrics to metadata
merged_seu <- AddMetaData(
  merged_seu,
  metadata = PercentageFeatureSet(
    merged_seu,
    features = mito_genes[mito_genes %in% rownames(merged_seu)]
  ),
  col.name = "percent_mito"
)

merged_seu <- AddMetaData(
  merged_seu,
  metadata = PercentageFeatureSet(
    merged_seu,
    features = ribo_genes[ribo_genes %in% rownames(merged_seu)]
  ),
  col.name = "percent_ribo"
)

# Add sequencing complexity metric
merged_seu$log10_genes_per_umi <- log10(merged_seu$nFeature_RNA) / log10(merged_seu$nCount_RNA)

cat("QC metrics calculated\n")
```

## Visualize QC metrics before filtering

```{r qc_violin, fig.height=10, fig.width=12}
# Set sample identity for visualization
Idents(merged_seu) <- "orig.ident"

# Violin plot of QC metrics
VlnPlot(
  merged_seu,
  features = c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo"),
  ncol = 2,
  pt.size = 0
)
```

## QC filtering

Apply standard QC thresholds:

```{r qc_filter}
# Store initial cell count
n_cells_initial <- ncol(merged_seu)

# Define QC thresholds
qc_thresholds <- list(
  nFeature_RNA = c(min = 400, max = 6000),
  nCount_RNA = c(min = 800, max = 40000),
  percent_mito = c(min = 0, max = 20),
  percent_ribo = c(min = 0, max = 60),
  log10_genes_per_umi = 0.7
)

# Apply filtering
merged_seu <- subset(
  merged_seu,
  subset = 
    nFeature_RNA >= qc_thresholds$nFeature_RNA["min"] &
    nFeature_RNA < qc_thresholds$nFeature_RNA["max"] &
    nCount_RNA >= qc_thresholds$nCount_RNA["min"] &
    nCount_RNA < qc_thresholds$nCount_RNA["max"] &
    percent_mito >= qc_thresholds$percent_mito["min"] &
    percent_mito < qc_thresholds$percent_mito["max"] &
    percent_ribo >= qc_thresholds$percent_ribo["min"] &
    percent_ribo < qc_thresholds$percent_ribo["max"] &
    log10_genes_per_umi > qc_thresholds$log10_genes_per_umi
)

n_cells_filtered <- ncol(merged_seu)
pct_retained <- 100 * n_cells_filtered / n_cells_initial

cat("QC filtering complete\n")
cat("Initial cells:", n_cells_initial, "\n")
cat("Filtered cells:", n_cells_filtered, "\n")
cat("Retained:", sprintf("%.1f%%", pct_retained), "\n")
```

## Remove small samples

```{r remove_small_samples}
# Remove samples with fewer than minimum cells
tab <- table(merged_seu$orig.ident)
keep_samples <- names(tab)[tab > min_cells_per_sample]

n_samples_before <- length(unique(merged_seu$orig.ident))
merged_seu <- subset(merged_seu, orig.ident %in% keep_samples)
n_samples_after <- length(unique(merged_seu$orig.ident))

cat("Samples before:", n_samples_before, "\n")
cat("Samples after (removed small samples):", n_samples_after, "\n")
cat("Total cells:", ncol(merged_seu), "\n")
```

## Normalize data

```{r normalize_data}
merged_seu <- NormalizeData(merged_seu)
cat("Data normalized\n")
```

---

# 6 - Save processed object

```{r save_object}
# Save the merged, filtered, standardized object
output_file <- file.path(data_dir, paste0(ds_name, "_whole.rds"))
saveRDS(merged_seu, output_file)
cat("Processed Zheng object saved to:", output_file, "\n")
cat("Final object dimensions:", nrow(merged_seu), "genes x", ncol(merged_seu), "cells\n")
```

---

## Session info

```{r session_info}
sessionInfo()
```

