---
title: "Oh et al. human bladder cancer - CD4 Data Processing"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script downloads, loads, and processes the Oh et al. 2020 bladder cancer CD4+ T cell dataset
for integration with the KLRG1 CD4-CTL reproducibility project. The script:

## Dataset Information
- **title**: Intratumoral CD4+ T Cells Mediate Anti-tumor Cytotoxicity in Human Bladder Cancer
- **doi**: https://doi.org/10.1016/j.cell.2020.05.017
- **Data source**: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149652
- **Output**: `scRNA-seq/cache/Oh_2020_32497499_whole.rds` (CD4+, high-quality cells)

---

# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(SingleCellExperiment)
library(STACAS)
library(SignatuR)
library(openxlsx)


read_excel_allsheets <- function(filename) {
  sheets <- openxlsx::getSheetNames(filename)
  x <- lapply(sheets, function(sheet_name) {
    openxlsx::read.xlsx(filename, sheet = sheet_name)
  })
  names(x) <- sheets
  x
}


options(timeout = 4000) # increase timeout for downloading
set.seed(321)
```

## Define parameters

```{r define_parameters}
# Dataset identifiers
ds_name <- "Oh_2020_32497499"
dataset_id <- "OhD"
species <- "human"

# QC filtering parameters
min_cells_per_sample <- 100
```

## Set up data paths

```{r setup_paths}
# Reference gene annotations (for gene symbol standardization) from STACAS
data("EnsemblGeneTable.Hs")

# Data paths
data_dir <- here("data")
input_data_path <- file.path(data_dir, "GSE149652")
list.files(input_data_path)
```

# 1 - Download data

Data downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149652


# 2 - Load data

## Load data

```{r load_raw_data}
# annotation column
annotation <- "cell_types"

cellinfo <- read.csv(file.path(input_data_path,
                               "GSE149652_CD4_TIL_droplet_cellinfo_matrice.csv.gz")) %>% 
  tibble::column_to_rownames("index")
# only for CD4 T cells
raw_counts <- data.table::fread(
  file.path(input_data_path, "GSE149652_CD4_TIL_droplet_count_matrice.csv.gz"),
  data.table = FALSE
)
rownames(raw_counts) <- raw_counts[[1]]
raw_counts <- raw_counts[, -1, drop = FALSE]

head(rownames(cellinfo))
head(colnames(raw_counts))

identical(rownames(cellinfo), colnames(raw_counts))
```


```{r create_Seurat}
seu <- CreateSeuratObject(counts = raw_counts,
                          meta.data = cellinfo)
# Standardize gene symbols
seu <- STACAS::StandardizeGeneSymbols(seu,
                                      slot = "counts",
                                      EnsemblGeneTable = EnsemblGeneTable.Hs)
```

# 3 - Quality Control and Filtering

## Calculate QC metrics

```{r calculate_qc_metrics}
library(SignatuR)
data("SignatuR")

# Get mitochondrial and ribosomal gene signatures
mito_genes <- GetSignature(SignatuR$Hs$Compartments$Mito) %>% 
  unlist()
ribo_genes <- GetSignature(SignatuR$Hs$Compartments$Ribo) %>%
  unlist()

# Add QC metrics to metadata
seu <- AddMetaData(
  seu,
  metadata = PercentageFeatureSet(
    seu,
    features = mito_genes[mito_genes %in% rownames(seu)]
  ),
  col.name = "percent_mito"
)

seu <- AddMetaData(
  seu,
  metadata = PercentageFeatureSet(
    seu,
    features = ribo_genes[ribo_genes %in% rownames(seu)]
  ),
  col.name = "percent_ribo"
)

# Add sequencing complexity metric
seu$log10_genes_per_umi <- log10(seu$nFeature_RNA) / log10(seu$nCount_RNA)

cat("QC metrics calculated\n")
```

## Visualize QC metrics before filtering
```{r qc_thresholds}
qc_thresholds <- list(
  nFeature_RNA = c(min = 200, max = 2000),
  nCount_RNA = c(min = 600, max = 3000),
  percent_mito = c(min = 0, max = 5),
  percent_ribo = c(min = 5, max = 40),
  log10_genes_per_umi = 0.7
)

## QC metrics distribution
qc_plot <- function(metadata, #seurat object metadata
                    var, # QC variables to plot
                    cutoffs,  # list of cutoffs with same names as QC vars
                    fill = NULL, # groupin variable for density plots
                    show.legend = T,
                    split = F
                    ) {
  

  metadata %>%
    ggplot(aes(.data[[var]])) +
    {if(is.null(fill)){
      geom_density(fill = "grey50",
                   color = "grey30")
    } else {
      geom_density(aes(fill = .data[[fill]]),
                   color = "grey10",
                   alpha = 0.5,
                   show.legend = show.legend)
    }}+
    {if(max(metadata[[var]])>100){
          scale_x_log10()}} +
    {if(!all(is.na(cutoffs[[var]]))){
      geom_vline(xintercept = cutoffs[[var]],
                 color = "red",
                 linetype="dashed")}} +
    {if(split){
      facet_wrap(~ .data[[fill]])}} +
    theme_bw() +
    theme(aspect.ratio = 1)
}
```


```{r qc_violin}
qclist <- lapply(names(qc_thresholds), function(c){  # names of QC metrics to evaluate
  qc_plot(seu@meta.data, var=c, cutoffs = qc_thresholds)
})
qclist
```

## QC filtering

Apply standard QC thresholds:

```{r qc_filter}
# Store initial cell count
n_cells_initial <- ncol(seu)

# Apply filtering
seu <- subset(
  seu,
  subset = 
    nFeature_RNA >= qc_thresholds$nFeature_RNA["min"] &
    nFeature_RNA < qc_thresholds$nFeature_RNA["max"] &
    nCount_RNA >= qc_thresholds$nCount_RNA["min"] &
    nCount_RNA < qc_thresholds$nCount_RNA["max"] &
    percent_mito >= qc_thresholds$percent_mito["min"] &
    percent_mito < qc_thresholds$percent_mito["max"] &
    percent_ribo >= qc_thresholds$percent_ribo["min"] &
    percent_ribo < qc_thresholds$percent_ribo["max"] &
    log10_genes_per_umi > qc_thresholds$log10_genes_per_umi
)

n_cells_filtered <- ncol(seu)
pct_retained <- 100 * n_cells_filtered / n_cells_initial

cat("QC filtering complete\n")
cat("Initial cells:", n_cells_initial, "\n")
cat("Filtered cells:", n_cells_filtered, "\n")
cat("Retained:", sprintf("%.1f%%", pct_retained), "\n")
```

## Remove small samples

```{r remove_small_samples}
# Remove samples with fewer than minimum cells
sample_id <- "Sample_ID"
tab <- table(seu@meta.data[[sample_id]])
keep_samples <- names(tab)[tab > min_cells_per_sample]

n_samples_before <- length(unique(seu@meta.data[[sample_id]]))
seu <- subset(seu, Sample_ID %in% keep_samples)
n_samples_after <- length(unique(seu@meta.data[[sample_id]]))

cat("Samples before:", n_samples_before, "\n")
cat("Samples after (removed small samples):", n_samples_after, "\n")
cat("Total cells:", ncol(seu), "\n")
```

## Normalize data

```{r normalize_data}
seu <- NormalizeData(seu)
cat("Data normalized\n")
```

---

# Save processed object

```{r save_object}
# Save the merged, filtered, standardized object
output_file <- file.path("cache", paste0(ds_name, ".rds"))
saveRDS(seu, output_file)
cat("Processed Oh object saved to:", output_file, "\n")
cat("Final object dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
```

---

## Session info

```{r session_info}
sessionInfo()
```

