---
title: "Kwek et al. human bladder cancer - Data Processing"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This script downloads, loads, and processes the Kwek et al. 2025 bladder cancer CD4+ T cell dataset
for integration with the KLRG1 CD4-CTL reproducibility project. The script:

## Dataset Information
- **title**: Identification and regulation of circulating tumor-TCR-matched cytotoxic CD4+ lymphocytes by KLRG1 in bladder cancer
- **doi**: https://insight.jci.org/articles/view/177373
- **Data source**: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293860
  Additionally, cell-level information was provided by David Oh (Corresponding) via email in file `cell_calls_annotations_josep_sept2026.csv`:
  
  ```text 
  please find attached a csv with the following:
  10X barcode ("cell_id")
  final cell annotation in object from JCI insight paper ("cell_calls")
  sample name ("title")
  the specific GEO accession that the 10X barcode refers to ("GEO")
  
  total 157,054 filtered single cells
  
  note that because individual samples were run as individual 10X reactions, there are overlapping 10X barcodes between samples. hence the GEO reference is necessary for you to know, which 10X barcode unambiguously belongs to which named GEO accession (and the cellranger outputs for that sample)
  
  good luck!
  
  dave
  ```
- **Output**: `scRNA-seq/cache/Kwek_2025_39655805.rds` (CD4+, high-quality cells)

---

# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(SingleCellExperiment)
library(STACAS)
library(SignatuR)
library(openxlsx)
library(BiocParallel)


read_excel_allsheets <- function(filename) {
  sheets <- openxlsx::getSheetNames(filename)
  x <- lapply(sheets, function(sheet_name) {
    openxlsx::read.xlsx(filename, sheet = sheet_name)
  })
  names(x) <- sheets
  x
}


options(timeout = 4000) # increase timeout for downloading
set.seed(321)
```

## Define parameters

```{r define_parameters}
# Dataset identifiers
ds_name <- "Kwek_2025_39655805"
dataset_id <- "KwekS"
species <- "human"

# QC filtering parameters
min_cells_per_sample <- 100
```

## Set up data paths

```{r setup_paths}
# Reference gene annotations (for gene symbol standardization) from STACAS
data("EnsemblGeneTable.Hs")

# Data paths
data_dir <- here("data")
input_data_path <- file.path(data_dir, "40299576")
list.files(input_data_path)
```

# 1 - Download data

Data downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293860 and cell-level metadata (including cell type annotation) provided by corresponding author (see Dataset Information above)


# 2 - Load data

## Load data

```{r load_raw_data}
# annotation column
annotation <- "cell_types"

cellinfo <- read.csv(file.path(input_data_path,
                               "cell_calls_annotations_josep_sept2026.csv")) %>% 
  # rename annotation col for consistency with other datasets
  dplyr::rename("cell_types" = "cell_calls",
                "original_cell_id" = "cell_id") %>% 
  # custom cell id and sample id
  mutate(title = gsub("antipdl1", "anti_pdl1", title),
        sample_id = paste(GEO, title,sep = "_"),
        cell_id = paste(sample_id, original_cell_id, sep = "_")) %>% 
  # filter only cd4 samples
  filter(grepl("cd4", title))

sample_md <- cellinfo %>% distinct(title, GEO, sample_id)

# only for CD4 T cells
cd4_files <- list.files(file.path(input_data_path, "GSE293860_RAW"),
                        pattern = "cd4",
                        full.names = T) %>% 
            # remove tcr sequencing files
            grep("tcr", ., invert = T, value = T)


```

## Run a loop to load all matrices for cd4

```{r}
library(Matrix)
samples <- sample_md$sample_id

seu_list <- bplapply(samples, 
                     BPPARAM = MulticoreParam(workers = 6, progressbar = T),
                     function(s){
                       files <- grep(s, cd4_files, value = T)
                       # load matrix
                       barcode_file <- files[grepl("barcodes.tsv.gz$", files)]
                       gene_file    <- files[grepl("genes.tsv.gz$", files)]
                       matrix_file  <- files[grepl("matrix.mtx.gz$", files)]
                       # Load sparse matrix
                       mat <- readMM(gzfile(matrix_file))
                       # Load genes and barcodes
                       genes <- read.delim(
                         gzfile(gene_file),
                         header = FALSE,
                         stringsAsFactors = FALSE
                       )
                       barcodes <- readLines(gzfile(barcode_file))
                       # Use gene symbols (column 2) or gene IDs (column 1)
                       gene_names <- genes[[1]]
                       # Assign cell and gene names
                       colnames(mat) <- paste(s, barcodes, sep = "_")
                       rownames(mat) <- gene_names
                       
                       # Create Seurat object
                       subset_md <- cellinfo %>% 
                         filter(sample_id == s) %>% 
                         tibble::column_to_rownames("cell_id")
                       mat <- mat[,rownames(subset_md)]
                       
                       seu <- CreateSeuratObject(counts = mat,
                                                 meta.data = subset_md)
                       return(seu)
                     }) 

```

## Create Seurat object

```{r merge_seurats}
seu <- merge(seu_list[[1]], seu_list[-1])
rm(seu_list); gc()
seu <- JoinLayers(seu)
```


## Standardize genes

```{r gene_Standard}
# Standardize gene symbols
seu <- STACAS::StandardizeGeneSymbols(seu,
                                      slot = "counts",
                                      EnsemblGeneTable = EnsemblGeneTable.Hs)
```

# 3 - Quality Control and Filtering

## Calculate QC metrics

```{r calculate_qc_metrics}
library(SignatuR)
data("SignatuR")

# Get mitochondrial and ribosomal gene signatures
mito_genes <- GetSignature(SignatuR$Hs$Compartments$Mito) %>% 
  unlist()
ribo_genes <- GetSignature(SignatuR$Hs$Compartments$Ribo) %>%
  unlist()

# Add QC metrics to metadata
seu <- AddMetaData(
  seu,
  metadata = PercentageFeatureSet(
    seu,
    features = mito_genes[mito_genes %in% rownames(seu)]
  ),
  col.name = "percent_mito"
)

seu <- AddMetaData(
  seu,
  metadata = PercentageFeatureSet(
    seu,
    features = ribo_genes[ribo_genes %in% rownames(seu)]
  ),
  col.name = "percent_ribo"
)

# Add sequencing complexity metric
seu$log10_genes_per_umi <- log10(seu$nFeature_RNA) / log10(seu$nCount_RNA)

cat("QC metrics calculated\n")
```

## Visualize QC metrics before filtering
```{r qc_thresholds}
qc_thresholds <- list(
  nFeature_RNA = c(min = 200, max = 2500),
  nCount_RNA = c(min = 300, max = 9000),
  percent_mito = c(min = 0, max = 7),
  percent_ribo = c(min = 5, max = 65),
  log10_genes_per_umi = 0.7
)

## QC metrics distribution
qc_plot <- function(metadata, #seurat object metadata
                    var, # QC variables to plot
                    cutoffs,  # list of cutoffs with same names as QC vars
                    fill = NULL, # groupin variable for density plots
                    show.legend = T,
                    split = F
                    ) {
  

  metadata %>%
    ggplot(aes(.data[[var]])) +
    {if(is.null(fill)){
      geom_density(fill = "grey50",
                   color = "grey30")
    } else {
      geom_density(aes(fill = .data[[fill]]),
                   color = "grey10",
                   alpha = 0.5,
                   show.legend = show.legend)
    }}+
    {if(max(metadata[[var]])>100){
          scale_x_log10()}} +
    {if(!all(is.na(cutoffs[[var]]))){
      geom_vline(xintercept = cutoffs[[var]],
                 color = "red",
                 linetype="dashed")}} +
    {if(split){
      facet_wrap(~ .data[[fill]])}} +
    theme_bw() +
    theme(aspect.ratio = 1)
}
```


```{r qc_violin}
qclist <- lapply(names(qc_thresholds), function(c){  # names of QC metrics to evaluate
  qc_plot(seu@meta.data, var=c, cutoffs = qc_thresholds)
})
qclist
```

## QC filtering

Apply standard QC thresholds:

```{r qc_filter}
# Store initial cell count
n_cells_initial <- ncol(seu)

# Apply filtering
seu <- subset(
  seu,
  subset = 
    nFeature_RNA >= qc_thresholds$nFeature_RNA["min"] &
    nFeature_RNA < qc_thresholds$nFeature_RNA["max"] &
    nCount_RNA >= qc_thresholds$nCount_RNA["min"] &
    nCount_RNA < qc_thresholds$nCount_RNA["max"] &
    percent_mito >= qc_thresholds$percent_mito["min"] &
    percent_mito < qc_thresholds$percent_mito["max"] &
    percent_ribo >= qc_thresholds$percent_ribo["min"] &
    percent_ribo < qc_thresholds$percent_ribo["max"] &
    log10_genes_per_umi > qc_thresholds$log10_genes_per_umi
)

n_cells_filtered <- ncol(seu)
pct_retained <- 100 * n_cells_filtered / n_cells_initial

cat("QC filtering complete\n")
cat("Initial cells:", n_cells_initial, "\n")
cat("Filtered cells:", n_cells_filtered, "\n")
cat("Retained:", sprintf("%.1f%%", pct_retained), "\n")
```

## Remove small samples

```{r remove_small_samples}
# Remove samples with fewer than minimum cells
seu@meta.data <- seu@meta.data %>% 
  dplyr::rename("Sample_ID" = "sample_id")
sample_id <- "Sample_ID"
tab <- table(seu@meta.data[[sample_id]])
keep_samples <- names(tab)[tab > min_cells_per_sample]

n_samples_before <- length(unique(seu@meta.data[[sample_id]]))
seu <- subset(seu, Sample_ID %in% keep_samples)
n_samples_after <- length(unique(seu@meta.data[[sample_id]]))

cat("Samples before:", n_samples_before, "\n")
cat("Samples after (removed small samples):", n_samples_after, "\n")
cat("Total cells:", ncol(seu), "\n")
```

## Normalize data

```{r normalize_data}
seu <- NormalizeData(seu)
cat("Data normalized\n")
```

---

# Save processed object

```{r save_object}
# Save the merged, filtered, standardized object
output_file <- file.path("cache", paste0(ds_name, ".rds"))
saveRDS(seu, output_file)
cat("Processed Oh object saved to:", output_file, "\n")
cat("Final object dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
```

---

## Session info

```{r session_info}
sessionInfo()
```

