---
title: "3 — DEG analysis: PBMC KLRG1+ vs KLRG1− CD4 T cells"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

## Input

- `scRNA-seq/cache/cd4_total_integrated_manual.rds` — Seurat object produced by
  `scRNA-seq/1.1.Integration_annotation.Rmd`. Contains all integrated CD4 T cells
  with columns `patient`, `tissue`, `annotation`, and `KLRG1_expression`
  (KLRG1+/KLRG1−) added in section 9.3 of that script.

## Output

- `scRNA-seq/cache/DEG_Pseudobulk_PBMC_KLRG1_POS_vs_KLRG1_NEG.rds` — DESeq2
  results data frame

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 7
)
```

## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(Seurat)
library(tidyverse)
library(DESeq2)
library(ggrepel)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(STRINGdb)
library(igraph)
library(ggraph)
library(here)
```

```{r paths}
cache_dir <- here::here("scRNA-seq", "cache")
in_rds    <- file.path(cache_dir, "cd4_total_integrated_manual.rds")

if (!file.exists(in_rds)) {
  stop("Input not found: ", in_rds,
       "\nRun scRNA-seq/1.1.Integration_annotation.Rmd first.")
}
```

---

# 1 — Load data

```{r load-data}
int <- readRDS(in_rds)
DefaultAssay(int) <- "RNA"

cat("Cells loaded:", ncol(int), "\n")
cat("Patients:    ", n_distinct(int$patient), "\n")
print(table(int$tissue, useNA = "ifany"))

# Ensure KLRG1_expression is present (added in 1.1 section 9.3)
if (!"KLRG1_expression" %in% colnames(int@meta.data)) {
  stop(
    "Column 'KLRG1_expression' not found in the loaded object.\n",
    "Re-run 1.1.Integration_annotation.Rmd (section 9.3) to add it."
  )
}
```

---

# 2 — PBMC subset

```{r pbmc-subset}
cd4.pbmc <- int[, int$tissue == "PBMC"]
cat("PBMC cells:", ncol(cd4.pbmc), "\n")
print(table(cd4.pbmc$KLRG1_expression, cd4.pbmc$patient))
```

## Cell-count bar chart per patient

```{r pbmc-cell-counts, fig.width=10, fig.height=5}
props <- cd4.pbmc@meta.data %>%
  dplyr::count(patient, KLRG1_expression) %>%
  dplyr::rename(Number_cells = n) %>%
  mutate(KLRG1_expression = factor(KLRG1_expression,
                                   levels = c("KLRG1+", "KLRG1-")))

ggplot(props, aes(patient, Number_cells, fill = KLRG1_expression)) +
  geom_col() +
  scale_fill_manual(values = c("KLRG1+" = "#0072B2", "KLRG1-" = "#D55E00")) +
  labs(
    x = "Patient",
    y = "Number of cells",
    fill = "KLRG1 status",
    title = "PBMC CD4 T cells — KLRG1+ vs KLRG1− per patient"
  ) +
  ggpubr::theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 3 — Pseudobulk aggregation

```{r pseudobulk-aggregate}
# Remove patients with very few cells in either compartment
min_cells_thresh <- 10
keep_patients <- cd4.pbmc@meta.data %>%
  dplyr::count(patient, KLRG1_expression) %>%
  group_by(patient) %>%
  filter(all(n >= min_cells_thresh)) %>%
  pull(patient) %>%
  unique()

cat("Patients retained for pseudobulk:", length(keep_patients), "\n")
cat(paste(" -", keep_patients, collapse = "\n"), "\n")

cd4.pbmc.filt <- cd4.pbmc[, cd4.pbmc$patient %in% keep_patients]

bu <- AggregateExpression(
  cd4.pbmc.filt,
  assays = "RNA",
  slot   = "counts",
  group.by = c("patient", "KLRG1_expression")
)[["RNA"]]

cat("Pseudobulk matrix dimensions:", nrow(bu), "genes ×", ncol(bu), "samples\n")
```

---

# 4 — DESeq2

## Build DESeq2 dataset

```{r deseq2-build}
# Remove TCR variable/diversity/joining genes (patient-specific)
trs <- grep("^TR[ABGD][VDJ]", rownames(bu), value = TRUE)
bu  <- bu[!rownames(bu) %in% trs, ]

# Build sample metadata from aggregated column names
# Column format: <patient>_<KLRG1_expression>
# KLRG1+ → KLRG1_POS ; KLRG1- → KLRG1_NEG (DESeq2 forbids special characters)
metadata <- data.frame(
  row.names = colnames(bu),
  Patient = sub("_KLRG1.*$", "", colnames(bu)),
  KLRG1_expression = sub("^.*_(KLRG1.*)$", "\\1", colnames(bu))
) %>%
  mutate(
    KLRG1_expression = gsub("[+]", "_POS", KLRG1_expression),
    KLRG1_expression = gsub("-",   "_NEG", KLRG1_expression),
    KLRG1_expression = factor(KLRG1_expression,
                              levels = c("KLRG1_NEG", "KLRG1_POS"))
  )

print(metadata)

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = bu,
  colData   = metadata,
  design    = ~ Patient + KLRG1_expression
)

# Remove genes with very low counts
dds <- dds[rowSums(counts(dds)) > 20, ]
cat("Genes after low-count filter:", nrow(dds), "\n")
```

## Quality control: PCA

```{r deseq2-pca}
dds <- estimateSizeFactors(dds)
vsd <- vst(dds, blind = TRUE)
plotPCA(vsd, intgroup = "KLRG1_expression") +
  ggtitle("PCA — pseudobulk PBMC KLRG1+/−")
```

## Differential expression

```{r deseq2-run}
dds <- DESeq(dds)
print(resultsNames(dds))

# Robustly select the KLRG1_POS vs KLRG1_NEG coefficient
klrg1_coef <- grep("KLRG1_POS", resultsNames(dds), value = TRUE)
cat("Using coefficient:", klrg1_coef, "\n")

res <- lfcShrink(dds, coef = klrg1_coef, type = "apeglm")

res_df <- as.data.frame(res) %>%
  tibble::rownames_to_column("gene") %>%
  arrange(padj)

cat("Significant DEGs (|log2FC| > log2(1.5), padj < 0.05):",
    sum(abs(res_df$log2FoldChange) >= log2(1.5) & res_df$padj < 0.05,
        na.rm = TRUE), "\n")

head(res_df, 20)
```

---

# 5 — Volcano plot: PBMC KLRG1+ vs KLRG1−

Corresponding to Figure 2B

```{r volcano, fig.width=13, fig.height=10}
fc  <- 1.5
fdr <- 0.05

pseudolog10 <- function(x) {
  sign(x) * log10(abs(x) + 1)
}

pseudolog10_breaks <- c(-16, -8, -4, -2, -1, 0, 1, 2, 4, 8, 16)
pseudolog10_labels <- as.character(pseudolog10_breaks)

cytotox_genes <- c("CST7", "GNLY", "NKG7", "GZMA", "GZMB", "GZMH", "GZMK", "PRF1")

vol_data <- res_df %>%
  mutate(
    Spec = ifelse(gene %in% cytotox_genes, "Relevant cytotoxic gene", "Other"),
    padj = ifelse(is.na(padj), 1, padj)
  ) %>%
  transmute(
    log2FC = log2FoldChange,
    sig    = -log10(padj),
    log2FC_pseudo = pseudolog10(log2FoldChange),
    sig_pseudo    = pseudolog10(-log10(padj)),
    gene,
    Spec
  )

set.seed(22)
vol_plot <- vol_data %>%
  ggplot(aes(x = log2FC_pseudo, y = sig_pseudo, color = Spec, label = gene)) +
  geom_point(size = 2, alpha = 0.95, show.legend = TRUE) +
  geom_vline(xintercept = pseudolog10(c(-log2(fc), log2(fc))),
             linetype = 2, linewidth = 0.3, col = "grey50") +
  geom_hline(yintercept = pseudolog10(-log10(fdr)),
             linetype = 2, linewidth = 0.3, col = "grey50") +
  ggrepel::geom_text_repel(
    data = subset(vol_data,
                  sig >= -log10(fdr) & abs(log2FC) >= log2(fc) &
                    Spec != "Relevant cytotoxic gene"),
    size = 4, fontface = "italic", max.overlaps = 5
  ) +
  ggrepel::geom_label_repel(
    aes(label = ifelse(Spec == "Relevant cytotoxic gene", gene, NA)),
    color = "white", fill = "#3B75AF", fontface = "italic",
    na.rm = TRUE
  ) +
  scale_color_manual(values = c("Other" = "grey30",
                                "Relevant cytotoxic gene" = "#3B75AF")) +
  scale_x_continuous(
    breaks = pseudolog10(pseudolog10_breaks),
    labels = pseudolog10_labels,
    name = "log2 Fold Change"
  ) +
  scale_y_continuous(
    breaks = pseudolog10(c(0, 1, 2, 4, 8, 16, 32, 64, 128, 256)),
    labels = c("0", "1", "2", "4", "8", "16", "32", "64", "128", "256"),
    name = "-log10 adj. p-value"
  ) +
  labs(
    color = "Gene type",
    title = "PBMC KLRG1+ vs KLRG1−  (pseudobulk DESeq2)"
  ) +
  ggpubr::theme_classic2() +
  theme(
    legend.title = element_text(size = 16, face = "bold"),
    legend.text  = element_text(size = 13),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 16, face = "bold")
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)))

vol_plot
```

---

# 6 — Export DEG table

```{r export-deg}
out_rds   <- file.path(cache_dir, "DEG_Pseudobulk_PBMC_KLRG1_POS_vs_KLRG1_NEG.rds")
out_excel <- file.path(cache_dir,  "DEG_Pseudobulk_PBMC_KLRG1_POS_vs_KLRG1_NEG.xlsx")

saveRDS(res_df, out_rds)
openxlsx::write.xlsx(
  x = res_df,
  file = out_excel,
  sheetName = "DEG",
  overwrite = TRUE
)

cat("DEG table saved:\n  ", out_rds, "\n  ", out_excel, "\n")
```

---

# 7 — GO enrichment (GOBP)

## Prepare gene lists

```{r go-prep}
de <- res_df

allEntrezs <- de$gene

selectedEntrezsUP <- de %>%
  filter(log2FoldChange > 0 & padj < 0.05) %>%
  pull(gene)

selectedEntrezsDOWN <- de %>%
  filter(log2FoldChange < 0 & padj < 0.05) %>%
  pull(gene)

# Signed -log10(p) metric for ranked enrichment
selectedEntrezs <- de %>%
  filter(!is.na(log2FoldChange)) %>%
  mutate(
    fcsign = sign(log2FoldChange),
    logP   = -log10(pvalue),
    metric = logP / fcsign
  ) %>%
  arrange(desc(metric)) %>%
  pull(metric, name = gene)

selectedEntrezs <- sort(selectedEntrezs, decreasing = TRUE)

cat("UP-regulated DEGs (padj < 0.05):  ", length(selectedEntrezsUP), "\n")
cat("DOWN-regulated DEGs (padj < 0.05):", length(selectedEntrezsDOWN), "\n")
```

## Load GOBP gene sets (MSigDB)

```{r go-load-msigdb}
m_df0 <- msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(gs_collection == "C5" & grepl("^GOBP_", gs_name)) %>%
  as.data.frame()

m_df <- m_df0[, c("gs_name", "gene_symbol")]

m_df2 <- m_df0 %>%
  distinct(gs_name, gs_description, gs_collection, gs_subcollection) %>%
  dplyr::rename(ID = gs_name)
```

## GO enrichment — upregulated genes

Corresponding to Figure 3A

Slight differences possible due to different package versions, but overall very similar results.

```{r go-up, fig.width=9, fig.height=9}
pth <- 0.05

ego <- enricher(
  selectedEntrezsUP,
  TERM2GENE      = m_df,
  universe       = allEntrezs,
  pAdjustMethod  = "fdr",
  pvalueCutoff   = pth
)

ego_res <- ego@result %>%
  filter(p.adjust < pth) %>%
  left_join(m_df2, by = "ID")

enrich_plot <- dotplot(ego,
                       showCategory = 20,
                       title = "UPREGULATED in KLRG1+ (PBMC)") +
  scale_color_gradientn(colours = c("#C65A5A", "#9C8797", "#2F6FA3")) +
  geom_point(alpha = 0.8, shape = 21, color = "black")

enrich_plot
```

---

# 8 — STRING interaction network

## Collect log2FC vector for node annotation

```{r}
lfc_vec <- de %>%
  dplyr::select(gene, log2FoldChange) %>%
  distinct() %>%
  tibble::deframe()
```


### Pick adhesion-related UP genes from GO enrichment
First, subset GO terms that are adhesion-related, then extract the contributing genes.
This ensures the network is functionally grounded in the GO results, not just arbitrary DE genes.

```{r}
# identify adhesion-related GO terms
adhesion_terms <- ego_res %>%
  filter(grepl("adhesion|cell adhesion|cell-cell", 
               Description, ignore.case = TRUE))

# extract genes contributing to those terms
adhesion_genes <- adhesion_terms %>%
  separate_rows(geneID, sep = "/") %>%
  distinct(geneID) %>%
  pull(geneID)

# keep only upregulated genes
adhesion_genes <- intersect(adhesion_genes, selectedEntrezsUP)

# add KLRG1 explicitly
genes_for_network <- unique(c("KLRG1", adhesion_genes))
```

### Query STRING for interaction data

```{r}
library(STRINGdb)

string_db <- STRINGdb$new(
  version = "11.5",
  species = 9606,
  score_threshold = 400,
  input_directory = ""
)

# map gene symbols to STRING IDs
mapped <- string_db$map(
  data.frame(gene = genes_for_network),
  "gene",
  removeUnmappedRows = TRUE
)

# retrieve interaction network
string_edges <- string_db$get_interactions(mapped$STRING_id)

# get STRING ID for KLRG1
klrg1_string_id <- mapped %>%
  filter(gene == "KLRG1") %>%
  pull(STRING_id)

# keep only edges directly connected to KLRG1
string_edges_klrg1 <- string_edges %>%
  filter(from == klrg1_string_id | to == klrg1_string_id)

connected_ids <- unique(c(
  string_edges_klrg1$from,
  string_edges_klrg1$to
))

mapped_klrg1 <- mapped %>%
  filter(STRING_id %in% connected_ids)
```

### Build graph object

```{r}
library(igraph)
valid_ids <- mapped_klrg1$STRING_id

edges <- string_edges_klrg1 %>%
  dplyr::filter(
    from %in% valid_ids &
    to   %in% valid_ids
  ) %>%
  dplyr::select(from, to, combined_score)

nodes <- mapped_klrg1 %>%
  transmute(
    name = STRING_id,      # IMPORTANT
    gene = gene,
    log2FC = lfc_vec[gene],
    is_klrg1 = gene == "KLRG1"
  )

g <- graph_from_data_frame(
  d = edges,
  vertices = nodes,
  directed = FALSE
)
```

### Render plot
string-like interaction map with KLRG1 highlighted.
combined_score is STRING’s overall confidence score for a protein–protein interaction.

```{r}
library(ggraph)
library(ggplot2)

set.seed(22)

gplot <- ggraph(g, layout = "fr") +

  ## edges: subtle, curved, confidence-weighted
  geom_edge_link(
    aes(width = combined_score),
    alpha = 0.35,
    color = "grey65",
    lineend = "round"
  ) +

  ## nodes: size = |log2FC|, KLRG1 emphasized
  geom_node_point(
    aes(
      size = abs(log2FC),
      fill = is_klrg1
    ),
    shape = 21,
    color = "black",
    stroke = 0.4
  ) +

  ## labels with halo for readability
  geom_node_text(
    aes(label = gene),
    repel = TRUE,
    size = 3.2,
    family = "sans",
    color = "black",
    bg.color = "white",
    bg.r = 0.15
  ) +

  ## scales
  scale_size_continuous(
    name = "|log2FC|",
    range = c(3, 10)
  ) +

  scale_fill_manual(
    values = c(`FALSE` = "#4C72B0",   # muted blue
               `TRUE`  = "#C44E52"),  # muted red
    labels = c("Adhesion-related genes", "KLRG1"),
    name = ""
  ) +

  scale_edge_width(
    range = c(0.3, 1.2),
    guide = "none"
  ) +

  ## theme tweaks
  theme_void() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 12,
      face = "bold"
    ),
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 9)
  ) +

  ggtitle("KLRG1-centered adhesion interaction network")

gplot
```

---

# Session information

```{r session-info}
sessionInfo()
```
