---
title: "scRNA-seq Custom Genes Expression — Cenerenti et al."
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

---

# Overview

This script loads the integrated/annotated Seurat object generated in
`1.2.Integration_annotation.Rmd` and measures the **IFN and MX1 expression**
using a pseudobulk per patient and tissue approach.

## Input

- `scRNA-seq/cache/cd4_total_integrated_manual.rds`

## Output

- Figures/tables are produced in the rendered report (no additional cache file is required).

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the notebook.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(Seurat)   # >= 4.0
library(Matrix)
library(ggplot2)
library(ggpubr)
library(tidyverse)
```

```{r paths}
# All outputs are written here; the directory already exists in the repo.
cache_dir <- here::here("scRNA-seq", "cache")
```

```{r helpers}
Log_Normalize <- function(mat,
                          total.counts,
                          scale.factor = 1e4,
                          margin = 2L) {
  stopifnot(inherits(mat, "dgCMatrix"))
  if (missing(total.counts) || is.null(total.counts)) {
    stop("`total.counts` must be provided")
  }

  margin <- as.integer(margin)
  if (!margin %in% c(1L, 2L)) stop("`margin` must be 1 (rows) or 2 (cols)")

  ids <- if (margin == 2L) colnames(mat) else rownames(mat)
  if (is.null(ids)) stop("Matrix must have row/col names for the chosen `margin`")

  total.counts <- total.counts[ids]
  if (anyNA(total.counts)) stop("`total.counts` is missing entries for some ids")
  if (any(total.counts <= 0)) stop("`total.counts` must be > 0 for all ids")

  if (margin == 2L) {
    mat <- Matrix::t(Matrix::t(mat) / total.counts)
  } else {
    mat <- mat / total.counts
  }

  mat@x <- log1p(mat@x * scale.factor)
  mat
}
```

# 1 — Load integrated object (from 1.2 script)

```{r load-integrated}
library(here)
int_path <- file.path(cache_dir, "cd4_total_integrated_manual.rds")

if (!file.exists(int_path)) {
  stop(
    "Missing integrated object: ", int_path,
    "\nRun scRNA-seq/1.2.Integration_annotation.Rmd first."
  )
}

int <- readRDS(int_path)
```

---

# 2 — IFN and MX1 expression

## Pseudobulk approach

```{r ifn-mx1-pseudobulk}
# Manual pseudobulk per (patient, tissue, annotation)
cgenes <- c("IFI6", "MX1")

cols <- c(
  "PBMC" = "#0072B2",
  "TILN" = "#D55E00"
)

DefaultAssay(int) <- "RNA"

# Expects these columns to exist:
# - ident:  annotation
# - sample: patient_tissue
if (!all(c("annotation", "patient", "tissue") %in% colnames(int@meta.data))) {
  stop(
    "Required metadata columns not found in int@meta.data. Need: annotation, patient, tissue.\n",
    "Available columns include: ", paste(head(colnames(int@meta.data), 25), collapse = ", ")
  )
}

int$patient_tissue <- paste(int$patient, int$tissue, sep = ".")

# Keep only groups with at least 5 cells
meta <- int@meta.data %>%
  dplyr::mutate(cell = rownames(int@meta.data))

groups <- meta %>%
  dplyr::count(patient, tissue, annotation, name = "n_cells") %>%
  dplyr::filter(n_cells >= 5) %>%
  dplyr::mutate(sample_id = paste(patient, tissue, annotation, sep = "_"))

meta <- meta %>%
  dplyr::inner_join(
    groups %>% dplyr::select(patient, tissue, annotation, n_cells, sample_id),
    by = c("patient", "tissue", "annotation")
  )

# Raw counts (all genes) for pseudobulk aggregation
counts <- Seurat::GetAssayData(int, assay = "RNA", layer = "counts")[, meta$cell, drop = FALSE]

# Pseudobulk sum per group
sample_index <- factor(meta$sample_id, levels = unique(groups$sample_id))
design <- Matrix::sparse.model.matrix(~ 0 + sample_index)
colnames(design) <- levels(sample_index)

mat_counts <- counts %*% design

# Log1p normalization (after pseudobulk aggregation)
total_counts <- Matrix::colSums(mat_counts)
mat_log1p <- Log_Normalize(
  mat = mat_counts,
  total.counts = total_counts,
  scale.factor = 1e4,
  margin = 2L
)

# Subset genes after log1p normalization
mat <- mat_log1p[cgenes, , drop = FALSE]
```

```{r ifn-mx1-tidy}
df <- mat %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("id") %>%
  pivot_longer(
    -id,
    names_to = "gene",
    values_to = "Expression Level"
  ) %>%
  separate(id, sep = "_", into = c("patient", "tissue", "annotation"))
```

```{r ifn-mx1-boxplot, fig.width=7, fig.height=4}
pos_box <- position_dodge2(width = 0.5, preserve = "single")
pos_points <- position_jitterdodge(
  jitter.width = 0.18,
  jitter.height = 0,
  dodge.width = 0.7
)

tissue_levels <- names(cols)
if (is.null(tissue_levels)) tissue_levels <- unique(df$tissue)

pl <- df %>%
  mutate(tissue = factor(tissue, levels = tissue_levels)) %>%
  ggplot(aes(x = annotation, y = `Expression Level`, fill = tissue)) +
  geom_boxplot(
    alpha = 0.2,
    outlier.color = NA,
    position = pos_box
  ) +
  geom_point(
    color = "black",
    shape = 21,
    position = pos_points,
    size = 2,
    alpha = 0.7,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = cols, drop = FALSE) +
  scale_color_manual(values = cols) +
  facet_wrap(~gene) +
  labs(x = "Identity") +
  ggpubr::theme_classic2() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank()
  )

pl
```

---

# Session information

```{r session-info}
sessionInfo()
```
