---
title: "Cross-dataset CTL CD4+ T cells comparison"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Context
In JEM revision, R3 in point 2 pointed:

*Many prior studies have also identified CD4 T cells with cytotoxic features (e.g. PMID: 41043413, PMID: 40299576, PMID: 39655805, and others). It would be valuable if the authors could compare the population identified in this study with previously described cytotoxic CD4 T cell subsets in cancer and other diseases, particularly in terms of their phenotypic and functional overlap.*

We tried to fetch these datasets but either data was not accessible or authors did not share annotation. We could only collect 2 datasets:

### 40299576
   - Actually proposed by reviewer
   - Paper: https://insight.jci.org/articles/view/177373
   Identification and regulation of circulating tumor-TCR-matched cytotoxic CD4+ lymphocytes by KLRG1 in bladder cancer
   - Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293860
   * Processed (cached): `2.3.Oh_processing.Rmd`
   
### 32497499
Proposed an extra dataset by Camila
Intratumoral CD4+ T Cells Mediate Anti-tumor Cytotoxicity in Human Bladder Cancer
https://doi.org/10.1016/j.cell.2020.05.017
- Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149652
* Processed (cached): `2.4.Kwek_processing.Rmd`


# Goals
- Signature overlap and label-transfer comparison against these published cytotoxic CD4 datasets.
- Add heatmap of conserved and unique markers.


# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(STACAS)
library(ProjecTILs)
library(SignatuR)
library(BiocParallel)
library(UCell)
```

# 1. Load processed datasets

These were harmonized, QC filtered and normalized in `2.3.Oh_processing.Rmd` and `2.4.Kwek_processing.Rmd`, respectively. 

In both, author's annotation are in `cell_types` and sample_id in `Sample_ID`.

```{r ident_sampleid}
author_annotation <- "cell_types"
sample_id <- "Sample_ID"
```

## Load Seurat into a list to run all code in loops
```{r load_seurat}
file_path <- c("cache/Oh_2020_32497499.rds",
               "cache/Kwek_2025_39655805.rds")
seu_list <- bplapply(file_path,
                     BPPARAM = MulticoreParam(workers = 2, progressbar = T),
                     function(f){
                        readRDS(f)
                     })
names(seu_list) <- tools::file_path_sans_ext(basename(file_path))  
```

What CD4T subtypes are annotated?

```{r}
lapply(seu_list, function(s){
  s@meta.data[[author_annotation]] %>% unique()
})
lapply(seu_list, function(s){
  s@meta.data[[sample_id]] %>% unique()
})
```

# 2. Annotate and project with inhouse data and ProjecTILs ref maps

## Load inhouse dataset and reference
```{r}
ref_paths <- c("cache/cd4_total_integrated_manual.rds", # produced in script 1.2.Integration_annotation.Rmd
               "cache/CD4T_human_ref_v2.rds")
ref <- bplapply(ref_paths,
                BPPARAM = MulticoreParam(workers = 2, progressbar = T),
                function(f){
                  readRDS(f)
                })
names(ref) <- c("inhouse", "ProjecTILs")

# inhouse dataset should be converted to projectils ref map for later classification
ref$inhouse <- make.reference(
  ref = ref$inhouse ,
  assay = "integrated",
  ndim = 50,
  seed = 22,
  recalculate.umap = FALSE,
  nfeatures = 1500,
  annotation.column = "annotation"
)
```

```{r}
lapply(ref, function(x){
  if("functional.cluster" %in% colnames(x@meta.data)){
    gb <- "functional.cluster"
  } else {
    gb <- "annotation"
  }
  DimPlot(x,
          group.by = gb)
})
```

## Annotate

```{r}
seu_list <- lapply(seu_list,
                   function(s){
                     
  s <- ProjecTILs.classifier(query = s,
                             ref = ref$inhouse,
                             filter.cells = F,
                             labels.col = "annotation",
                             split.by = sample_id,
                             ncores = 7)
  s$pred_inhouse <- s$annotation
  
  s <- ProjecTILs.classifier(query = s,
                             ref = ref$ProjecTILs,
                             filter.cells = F,
                             labels.col = "functional.cluster",
                             split.by = sample_id,
                             ncores = 7)
  s$pred_ProjecTILs <- s$functional.cluster
  
  return(s)
})
```


# 3. Plot confusion matrix plots
Show the agreement between author's annotation and the projection of our manual annotation, spepecially for inhouse dataset.



```{r confusion_plot, fig.width=10, fig.height=5}
source("utils/confusion_plot.R")
conf_plot <- lapply(names(seu_list), function(s){
  plot.confusion(object = seu_list[[s]],
                 var.1 = author_annotation,
                 var.2 = "pred_inhouse",
                 ylab = "Prediction",
                 xlab = "Author's annotation",
                 plot.title = s,
                 type = "tile",
                 relative = T)
})
conf_plot
```

# 4. Signatures

We reproduce heatmap done in `2.2.Annotatoin_multicancer_dataset.Rmd` and corresponding to Suppl Figure 1G.

Define the main cd4-ctl and thf markers

```{r define_markers}
cd4_cytotoxic_markers <- c("GZMB", "GNLY", "NKG7", "FGFBP2", "PRF1", "CST7", "GZMA",
                           "GZMH", "CTSW")
tfh_markers <- c("CXCR5", "PDCD1", "CXCL13", "IL21", "TOX")
feats_heatmap <- c(cd4_cytotoxic_markers, tfh_markers)
# define which cell types to fetch
sel_ct <- "cytotoxic|tfh|cxcl13|gzmb|gzmk"
sel_ct <- "cytotoxic|gzmb|gzmk"
```

## Join query datasets and inhouse dataset into the same object
```{r}
# add dataset of origin
seu_list <- lapply(names(seu_list), function(s){
  seu_list[[s]]$dataset <- s
  return(seu_list[[s]])
})
```

```{r}
ref$inhouse@meta.data[[sample_id]] <- ref$inhouse$MULTI_classification
ref$inhouse@meta.data[[author_annotation]] <- ref$inhouse$annotation
ref$inhouse@meta.data[["dataset"]] <- "ThiStudy"
```

```{r}
mer <- merge(ref$inhouse, seu_list)
# subset only for CTL and TFH-like cells
mer <- mer[,grepl(sel_ct, mer@meta.data[[author_annotation]], ignore.case = T)]
table(mer[[author_annotation]])
```
## Heatmap per sample

### Build pseudobulk

```{r}
# Raw counts
counts <- GetAssayData(mer, assay = "RNA", layer = "counts")
# Group cells by sample + cell type
group <- interaction(
  mer@meta.data[[sample_id]],
  mer@meta.data[[author_annotation]],
  mer@meta.data$dataset,
  sep = "--",
  drop = TRUE
)
# Pseudobulk: sum raw counts within each sample × cell type
counts_pb <- sapply(
  levels(group),
  function(g) {
    Matrix::rowSums(counts[, group == g, drop = FALSE])
  }
)
counts_pb <- Matrix::Matrix(counts_pb, sparse = TRUE)
# Clean group names
colnames(counts_pb) <- levels(group)
# Log-normalize
counts_pb_log <- log1p(
  t(t(counts_pb) / Matrix::colSums(counts_pb) * 1e4)
)

```


```{r}
library(pheatmap)

ann_col <- data.frame(id = colnames(counts_pb_log)) %>% 
  separate(
    id,
    sep = "--",
    into = c("sample_id", "cell_type", "dataset"),
    remove = FALSE
  ) %>% 
  select(-sample_id) %>% 
  arrange(desc(dataset), cell_type) %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = unique(cell_type)
    )
  ) %>%
  tibble::column_to_rownames("id")

mat <- counts_pb_log[cd4_cytotoxic_markers,rownames(ann_col)]

pheatmap(mat = mat,
         scale = "row",
         cluster_cols = F,
         cluster_rows = F,
         show_colnames = F, 
         annotation_col = ann_col
         )
```


# 5. Markers overlaps

### Findmarkers
```{r}
# get commone set of features in 3 seurat object
DefaultAssay(ref$inhouse) <- "RNA"
common_feats <- Reduce(
  intersect,
  list(
    Features(seu_list$Oh_2020_32497499),
    Features(seu_list$Kwek_2025_39655805),
    Features(ref$inhouse)
  )
)
# remove also tcr and hla genes
common_feats <- common_feats[
  !grepl(
    "^(TR[ABDG][VCJ]|HLA-)",
    common_feats
  )
]

top_genes <- 25
scmarkers <- lapply(c(seu_list, ref$inhouse),
  function(s) {
    # combine cytotoxic cells
    ctl_annot <- paste0("CD4-CTL_", unique(s$dataset))
    s$new_annot <- ifelse(grepl(sel_ct, s@meta.data[[author_annotation]], ignore.case = T),
                          ctl_annot, 
                          as.character(s@meta.data[[author_annotation]]))
    naive_annot <- grep("memory|CM", s@meta.data[[author_annotation]],
                        ignore.case = T, value = T) %>% unique()
    cat("Dataset: ", unique(s$dataset), "\n",
          "Comparing ", unique(grep("cytotoxic|gzmb",
                                    s@meta.data[[author_annotation]],
                                    ignore.case = T, value = T)),
          " vs ", naive_annot)
    
    Idents(s) <- "new_annot"
    m <- FindMarkers(
      s,
      ident.1 = ctl_annot,
      ident.2 = naive_annot,
      assay = "RNA",
      features = common_feats,
      test.use = "bimod",
      min.pct = 0.3,
      min.diff.pct = 0.1,
      min.cells.feature = 10,
      verbose = FALSE,
      only.pos = TRUE
    )
    
    m <- m %>% 
      arrange(desc(avg_log2FC)) %>% 
      head(top_genes) %>% 
      mutate(dataset = unique(s$dataset)) %>% 
      tibble::rownames_to_column("gene")
    
    return(m)
  }
) %>% 
  data.table::rbindlist()
```

```{r}
table(scmarkers$dataset) %>% sort()
```

## Upset plot

```{r}
marker_sets <- scmarkers %>%
  group_by(dataset) %>%
  summarise(
    genes = list(unique(gene)),
    .groups = "drop"
  )

marker_list <- setNames(marker_sets$genes, marker_sets$dataset)
```

```{r}
marker_df0 <- tibble(
  gene = unique(unlist(marker_list))
)

for (ct in names(marker_list)) {
  marker_df0[[ct]] <- marker_df0$gene %in% marker_list[[ct]]
}

marker_df0
```


```{r}
library(UpSetR)

# Convert list of gene sets to UpSetR input
marker_df <- fromList(marker_list)

# Plot overlap
up <- upset(
  marker_df,
  sets = names(marker_list),
  keep.order = TRUE,
  order.by = "freq",
  nsets = length(marker_list),
  nintersects = NA
)
up
```



## Euler's plot

```{r}
library(eulerr)

fit <- euler(marker_list)

plot(
  fit,
  quantities = list(
    fontsize = 10,
    col = "black"
  ),
  labels = list(
    fontsize = 11
  ),
  edges = TRUE
)
```







