---
title: "Cross-dataset CTL CD4+ T cells comparison"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Context
In JEM revision, R3 in point 2 pointed:

*Many prior studies have also identified CD4 T cells with cytotoxic features (e.g. PMID: 41043413, PMID: 40299576, PMID: 39655805, and others). It would be valuable if the authors could compare the population identified in this study with previously described cytotoxic CD4 T cell subsets in cancer and other diseases, particularly in terms of their phenotypic and functional overlap.*

We tried to fetch these datasets but either data was not accessible or authors did not share annotation. We could only collect 2 datasets:

### 40299576
   - Actually proposed by reviewer
   - Paper: https://insight.jci.org/articles/view/177373
   Identification and regulation of circulating tumor-TCR-matched cytotoxic CD4+ lymphocytes by KLRG1 in bladder cancer
   - Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293860
   * Processed (cached): `2.3.Oh_processing.Rmd`
   
### 32497499
Proposed an extra dataset by Camila
Intratumoral CD4+ T Cells Mediate Anti-tumor Cytotoxicity in Human Bladder Cancer
https://doi.org/10.1016/j.cell.2020.05.017
- Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149652
* Processed (cached): `2.4.Kwek_processing.Rmd`

### Chu et al., 2023
(https://www.nature.com/articles/s41591-023-02371-y) which identifies T-stressed cell subsets but also a CD4-CTL population

# Goals
- Signature overlap and label-transfer comparison against these published cytotoxic CD4 datasets.
- Add heatmap of conserved and unique markers.


# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(STACAS)
library(ProjecTILs)
library(SignatuR)
library(BiocParallel)
library(UCell)

source("utils/utils.R")
```

# 1. Load processed datasets

These were harmonized, QC filtered and normalized in `2.3.Oh_processing.Rmd` and `2.4.Kwek_processing.Rmd`, respectively. 

In both, author's annotation are in `cell_types` and sample_id in `Sample_ID`.

```{r ident_sampleid}
author_annotation <- "cell_types"
sample_id <- "Sample_ID"
```

## Load Seurat into a list to run all code in loops
```{r load_seurat}
file_path <- c("cache/Oh_2020_32497499.rds",
               "cache/Kwek_2025_39655805.rds",
               "cache/ChuY_2023_37248301.rds" # from ChuY_2023_37248301_Nofilt_whole.rds
) 
seu_list <- lapply(file_path,
                   function(f){
                     readRDS(f)
                   }) 
names(seu_list) <- tools::file_path_sans_ext(basename(file_path)) 

# harmonize ChuY dataset
## Keep only CD4
seu_list$ChuY_2023_37248301 <- seu_list$ChuY_2023_37248301$CD4
## choose author's annotation
seu_list$ChuY_2023_37248301@meta.data[[author_annotation]] <- seu_list$ChuY_2023_37248301@meta.data[["Cellsubtype"]]
# this dataset do not contain sample, we will have to use batch of origin
seu_list$ChuY_2023_37248301@meta.data[[sample_id]] <- seu_list$ChuY_2023_37248301@meta.data[["batch"]]
```



What CD4T subtypes are annotated?

```{r}
lapply(seu_list, function(s){
  s@meta.data[[author_annotation]] %>% unique()
})
lapply(seu_list, function(s){
  s@meta.data[[sample_id]] %>% unique()
})
```

# 2. Annotate and project with inhouse data and ProjecTILs ref maps

## Load inhouse dataset and reference
```{r}
ref_paths <- list("inhouse" = "cache/cd4_total_integrated_manual.rds" 
                  # produced in script 1.2.Integration_annotation.Rmd
               #"ProjecTILs" = cache/CD4T_human_ref_v2.rds")
)
ref <- bplapply(ref_paths,
                BPPARAM = MulticoreParam(workers = 2, progressbar = T),
                function(f){
                  readRDS(f)
                })

# inhouse dataset should be converted to projectils ref map for later classification
ref$inhouse <- make.reference(
  ref = ref$inhouse ,
  assay = "integrated",
  ndim = 50,
  seed = 22,
  recalculate.umap = FALSE,
  nfeatures = 1500,
  annotation.column = "annotation"
)

```

```{r}
lapply(ref, function(x){
  if("functional.cluster" %in% colnames(x@meta.data)){
    gb <- "functional.cluster"
  } else {
    gb <- "annotation"
  }
  DimPlot(x,
          group.by = gb)
})
```

## Annotate

```{r}
seu_list <- lapply(seu_list,
                   function(s){
                     
  s <- ProjecTILs.classifier(query = s,
                             ref = ref$inhouse,
                             filter.cells = F,
                             labels.col = "annotation",
                             split.by = sample_id,
                             ncores = 7)
  s$pred_inhouse <- s$annotation
  # 
  # s <- ProjecTILs.classifier(query = s,
  #                            ref = ref$ProjecTILs,
  #                            filter.cells = F,
  #                            labels.col = "functional.cluster",
  #                            split.by = sample_id,
  #                            ncores = 7)
  # s$pred_ProjecTILs <- s$functional.cluster
  
  return(s)
})
```


# 3. Plot confusion matrix plots
Show the agreement between author's annotation and the projection of our manual annotation, spepecially for inhouse dataset.



```{r confusion_plot, fig.width=10, fig.height=5}
source("utils/confusion_plot.R")
conf_plot <- lapply(names(seu_list), function(s){
  plot.confusion(object = seu_list[[s]],
                 var.1 = author_annotation,
                 var.2 = "pred_inhouse",
                 ylab = "Prediction",
                 xlab = "Author's annotation",
                 plot.title = s,
                 type = "tile",
                 relative = T)
})
conf_plot
```

# 4. Signatures

## 4.1 Gene expression

We reproduce heatmap done in `2.2.Annotatoin_multicancer_dataset.Rmd` and corresponding to Suppl Figure 1G.

Define the main cd4-ctl and thf markers

```{r define_markers}
cd4_cytotoxic_markers <- c("GZMB", "GNLY", "NKG7", "FGFBP2", "PRF1", "CST7", "GZMA",
                           "GZMH", "CTSW", "KLRG1")
tfh_markers <- c("CXCR5", "PDCD1", "CXCL13", "IL21", "TOX")
feats_heatmap <- c(cd4_cytotoxic_markers, tfh_markers)
# define which cell types to fetch
sel_ct <- "cytotoxic|tfh|cxcl13|gzmb|gzmk" # ctl and tfh
sel_ct <- "cytotoxic|gzmb|gzmk|ctl" # only ctl
```

## Join query datasets and inhouse dataset into the same object
```{r}
# add dataset of origin
seu_list <- lapply(names(seu_list), function(s){
  seu_list[[s]]$dataset <- s
  return(seu_list[[s]])
})
names(seu_list) <- tools::file_path_sans_ext(basename(file_path)) 
```

```{r}
ref$inhouse@meta.data[[sample_id]] <- ref$inhouse$MULTI_classification
ref$inhouse@meta.data[[author_annotation]] <- ref$inhouse$annotation
ref$inhouse@meta.data[["dataset"]] <- "ThiStudy"
ref$inhouse$treatment <- "Untreated"

# accomodate metadata
seu_list$Oh_2020_32497499$treatment <- factor(seu_list$Oh_2020_32497499$treatment,
                                              levels = c("Atezo", "Chemo", "None"),,
                                              labels = c("antiPDL1", "Chemo", "Untreated"))
seu_list$Kwek_2025_39655805$treatment <- case_when(
                                                   grepl("untreated", seu_list$Kwek_2025_39655805$title) ~ "Untreated",
                                                   grepl("chemo", seu_list$Kwek_2025_39655805$title) ~ "Chemo",
                                                   grepl("pdl1", seu_list$Kwek_2025_39655805$title) ~ "antiPDL1",
                                                   grepl("healthy", seu_list$Kwek_2025_39655805$title) ~ "Healthy"
                                                   )
seu_list$Kwek_2025_39655805$tissue <- case_when(
                                                   grepl("normal", seu_list$Kwek_2025_39655805$title) ~ "Normal",
                                                   grepl("blood", seu_list$Kwek_2025_39655805$title) ~ "PBMC",
                                                   grepl("tumor", seu_list$Kwek_2025_39655805$title) ~ "Tumor"
                                                   )
```

```{r}
allmer <- merge(ref$inhouse, seu_list)
DefaultAssay(allmer) <- "RNA"
# subset only for CTL and TFH-like cells
mer <- allmer[,grepl(sel_ct, allmer@meta.data[[author_annotation]], ignore.case = T)]
table(mer[[author_annotation]])
table(mer$tissue, mer$dataset, useNA = "ifany")
table(mer$treatment, mer$dataset, useNA = "ifany")
```
## Dotplot

```{r}
mer$ds_ct <- factor(paste(mer$dataset,
                          mer@meta.data[[author_annotation]],
                          sep = "|"),
                    levels = c("ThiStudy|Cytotoxic",
                               "Kwek_2025_39655805|GZMB+",
                               "Kwek_2025_39655805|GZMK+",
                               "Oh_2020_32497499|CD4GZMB",
                               "Oh_2020_32497499|CD4GZMK",
                               "ChuY_2023_37248301|Tctl"
                    ))
dt <- dotplot_seurat_style(
  object = mer,
  features = cd4_cytotoxic_markers,
  ident_col = "ds_ct",
  split_by = NULL,
  assay = "RNA",
  scale_expression = F
) + 
  theme(axis.text.y = element_text(size = 18),
        axis.title = element_blank(),
        title = element_text(size = 22)) +
  ggplot2::ggtitle("Cytotoxic marker expression")
dt
```

## Violin plot

```{r}
vln <- VlnPlot(mer,
               group.by = "ds_ct",
               features = cd4_cytotoxic_markers, 
               flip = T,
               stack = T,
               fill.by = "ident") +
  NoLegend()

vln  
```

### export figures

```{r}
new_fig <- "~/OneDrive - Université de Genève/CSI/LabPapers/Cenerenti_KLRG1_CTL/figures/new_figs_Sep26"

fig_path <- file.path(new_fig, "DotPlot_CTLmarkers_otherDatasets.pdf")
pdf(fig_path, width = 6.5, height = 8)
dt
dev.off()

fig_path <- file.path(new_fig, "ViolinPlot_CTLmarkers_otherDatasets.pdf")
pdf(fig_path, width = 4, height = 6)
vln
dev.off()
```



## Heatmap per sample

### Build pseudobulk

```{r}
# Raw counts
counts <- GetAssayData(mer, assay = "RNA", layer = "counts")
# Group cells by sample + cell type
group <- interaction(
  mer@meta.data[[sample_id]],
  mer@meta.data[[author_annotation]],
  mer@meta.data$dataset,
  sep = "--",
  drop = TRUE
)
# Pseudobulk: sum raw counts within each sample × cell type
counts_pb <- sapply(
  levels(group),
  function(g) {
    Matrix::rowSums(counts[, group == g, drop = FALSE])
  }
)
counts_pb <- Matrix::Matrix(counts_pb, sparse = TRUE)
# Clean group names
colnames(counts_pb) <- levels(group)
# Log-normalize
counts_pb_log <- log1p(
  t(t(counts_pb) / Matrix::colSums(counts_pb) * 1e4)
)

```


```{r}
library(pheatmap)

ann_col <- data.frame(id = colnames(counts_pb_log)) %>% 
  separate(
    id,
    sep = "--",
    into = c("sample_id", "cell_type", "dataset"),
    remove = FALSE
  ) %>% 
  select(-sample_id) %>% 
  arrange(desc(dataset), cell_type) %>%
  mutate(
    cell_type = factor(
      cell_type,
      levels = unique(cell_type)
    )
  ) %>%
  tibble::column_to_rownames("id")

mat <- counts_pb_log[cd4_cytotoxic_markers,rownames(ann_col)]

pheatmap(mat = mat,
         scale = "row",
         cluster_cols = F,
         cluster_rows = F,
         show_colnames = F, 
         annotation_col = ann_col,
         main = "Cytotoxic markers in CD4-T Cytotoxic-like populations\nPseudobulk per sample",
         )
```

## 4.2 Signature scoring

```{r}
DefaultAssay(allmer) <- "RNA"
DefaultAssay(ref$inhouse) <- "RNA"
# check if some dataset do not some gene of the signature
common_feats_ctl <- Reduce(
  intersect,
  list(
    Features(seu_list$Oh_2020_32497499),
    Features(seu_list$Kwek_2025_39655805),
    Features(seu_list$ChuY_2023_37248301),
    Features(ref$inhouse),
    cd4_cytotoxic_markers
  )
)

allmer <- AddModuleScore_UCell(allmer,
                            features = list(Cytotoxic = cd4_cytotoxic_markers),
                            name = "",
                            ncores = 7)
```

Tidy dataframe & plotting

```{r score_plot_all_ct}
udf <- allmer@meta.data %>% 
  group_by(.data[[sample_id]], .data[[author_annotation]],
         tissue, treatment, dataset) %>% 
  summarize(Cytotoxic_score = mean(Cytotoxic))
  
# plot
upl <- udf %>% 
  ggplot(aes(.data[[author_annotation]], Cytotoxic_score,
             fill = .data[[author_annotation]])) +
  geom_boxplot(
    outlier.color = NA,
    show.legend = FALSE,
    alpha = 0.4
  ) +
  geom_jitter(
    color = "grey22",
    show.legend = FALSE,
    shape = 21,
    width = 0.1,
    alpha = 0.8
  ) +
  labs(
    x = "Cell type annotation",
    y = "UCell score",
    title = "Gene signature scoring - Mean per sample"
  ) +
  facet_wrap(~dataset,
             scales = "free_x") +
  ggpubr::theme_classic2() +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

upl
```

```{r score_plot_ctl}
cdf <- allmer@meta.data %>% 
  filter(grepl(sel_ct, .data[[author_annotation]], ignore.case = T)) %>% 
  group_by(.data[[sample_id]], .data[[author_annotation]],
           tissue, treatment, dataset) %>% 
  summarize(Cytotoxic_score = mean(Cytotoxic), .groups = "drop") %>% 
  ungroup() %>% 
  mutate(cell_types = factor(.data[[author_annotation]],
                                             levels = c("Cytotoxic","CD4GZMB",
                                                        "CD4GZMK", "GZMB+","GZMK+",
                                                        "Tctl"))
         )

comps <- list(c("Cytotoxic", "CD4GZMB"),
              c("Cytotoxic", "CD4GZMK"),
              c("Cytotoxic", "GZMB+"),
              c("Cytotoxic", "GZMK+"),
              c("Cytotoxic", "Tctl")
              )
  
# plot
cpl <- cdf %>% 
  ggplot(aes(.data[[author_annotation]], Cytotoxic_score,
             fill = dataset)) +
  geom_boxplot(
    outlier.color = NA,
    alpha = 0.4
  ) +
  geom_jitter(
    color = "grey22",
    show.legend = FALSE,
    shape = 21,
    width = 0.1,
    alpha = 0.8
  ) +
  ggpubr::stat_compare_means(
    comparisons = comps,
    method = "wilcox.test",
    label = "p.format"
  ) +
  labs(
    x = "Cell type annotation",
    y = "UCell score",
    title = "Gene signature scoring of cytotoxic-like types",
    subtitle = "Mean per sample"
  ) +
  ggpubr::theme_classic2() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

cpl
```

```{r}
tpl <- cdf %>% 
  ggplot(aes(.data[[author_annotation]], Cytotoxic_score,
             fill = treatment)) +
  geom_boxplot(
    outlier.color = NA,
    alpha = 0.4,
    position = position_dodge(width = 0.8)
  ) +
  # geom_point(
  #   aes(group = tissue),
  #   color = "grey22",
  #   show.legend = FALSE,
  #   shape = 21,
  #   alpha = 0.8,
  #   position = position_jitterdodge(
  #     jitter.width = 0.1,
  #     dodge.width = 0.8
  #   )
  # ) +
  labs(
    x = "Cell type annotation",
    y = "UCell score",
    title = "Gene signature scoring of cytotoxic-like types",
    subtitle = "Mean per sample"
  ) +
  facet_wrap(~tissue) +
  ggpubr::theme_classic2() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

tpl
```





# 5. Markers overlaps

### Findmarkers
```{r}
# get commone set of features in 3 seurat object
DefaultAssay(ref$inhouse) <- "RNA"
common_feats <- Reduce(
  intersect,
  list(
    Features(seu_list$Oh_2020_32497499),
    Features(seu_list$Kwek_2025_39655805),
    Features(ref$inhouse)
  )
)
# remove also tcr and hla genes
common_feats <- common_feats[
  !grepl(
    "^(TR[ABDG][VCJ]|HLA-)",
    common_feats
  )
]

top_genes <- 25
scmarkers <- lapply(c(seu_list, ref$inhouse),
  function(s) {
    # combine cytotoxic cells
    ctl_annot <- paste0("CD4-CTL_", unique(s$dataset))
    s$new_annot <- ifelse(grepl(sel_ct, s@meta.data[[author_annotation]], ignore.case = T),
                          ctl_annot, 
                          as.character(s@meta.data[[author_annotation]]))
    naive_annot <- grep("memory|CM", s@meta.data[[author_annotation]],
                        ignore.case = T, value = T) %>% unique()
    cat("Dataset: ", unique(s$dataset), "\n",
          "Comparing ", unique(grep(sel_ct,
                                    s@meta.data[[author_annotation]],
                                    ignore.case = T, value = T)),
          " vs ", naive_annot, "\n")
    
    Idents(s) <- "new_annot"
    m <- FindMarkers(
      s,
      ident.1 = ctl_annot,
      ident.2 = naive_annot,
      assay = "RNA",
      features = common_feats,
      test.use = "bimod",
      min.pct = 0.3,
      min.diff.pct = 0.1,
      min.cells.feature = 10,
      verbose = FALSE,
      only.pos = TRUE
    )
    
    m <- m %>% 
      arrange(desc(avg_log2FC)) %>% 
      head(top_genes) %>% 
      mutate(dataset = unique(s$dataset)) %>% 
      tibble::rownames_to_column("gene")
    
    return(m)
  }
) %>% 
  data.table::rbindlist()
```

```{r}
table(scmarkers$dataset) %>% sort()
```

## Upset plot

```{r}
marker_sets <- scmarkers %>%
  group_by(dataset) %>%
  summarise(
    genes = list(unique(gene)),
    .groups = "drop"
  )

marker_list <- setNames(marker_sets$genes, marker_sets$dataset)
```

```{r}
# Unique markers for each set
unique_markers <- lapply(names(marker_list), function(x) {
  others <- setdiff(names(marker_list), x)
  setdiff(
    marker_list[[x]],
    unique(unlist(marker_list[others]))
  )
})
names(unique_markers) <- names(marker_list)

# Pairwise intersections
pairwise_intersections <- list(
  Kwek_Oh = intersect(
    marker_list$Kwek_2025_39655805,
    marker_list$Oh_2020_32497499
  ),
  Kwek_ThiStudy = intersect(
    marker_list$Kwek_2025_39655805,
    marker_list$ThiStudy
  ),
  Oh_ThiStudy = intersect(
    marker_list$Oh_2020_32497499,
    marker_list$ThiStudy
  )
)

# Intersection of all three
three_way_intersection <- Reduce(intersect, marker_list)

unique_markers
pairwise_intersections
three_way_intersection

clipr::write_clip(paste(pairwise_intersections$Oh_ThiStudy, collapse = ", "))
```


```{r}
marker_df0 <- tibble(
  gene = unique(unlist(marker_list))
)

for (ct in names(marker_list)) {
  marker_df0[[ct]] <- marker_df0$gene %in% marker_list[[ct]]
}

marker_df0
```


```{r}
library(UpSetR)

# Convert list of gene sets to UpSetR input
marker_df <- fromList(marker_list)

# Plot overlap
up <- upset(
  marker_df,
  sets = names(marker_list),
  keep.order = TRUE,
  order.by = "freq",
  nsets = length(marker_list),
  nintersects = NA
)
up
```



## Euler's plot

```{r}
library(eulerr)

fit <- euler(marker_list)

plot(
  fit,
  quantities = list(
    fontsize = 10,
    col = "black"
  ),
  labels = list(
    fontsize = 11
  ),
  edges = TRUE
)
```

# 6. scTypeEval

Try to use confusion/heatmap plot to evaluate correspondance of different annotations

```{r}
get_features_filtered <- function(seu,
                                  sel_ct = "cytotoxic|gzmb|gzmk",
                                  min_row = 50){
  seu <- seu[, grepl(sel_ct,
                     seu@meta.data[[author_annotation]],
                     ignore.case = TRUE)]
  
  counts <- Seurat::GetAssayData(seu, layer = "counts")
  keep_genes <- Matrix::rowSums(counts) > min_row
  
  rownames(counts)[keep_genes]
}
DefaultAssay(ref$inhouse) <- "RNA"

common_feats_ct <- Reduce(
  intersect,
  list(
    get_features_filtered(seu_list$Oh_2020_32497499),
    get_features_filtered(seu_list$Kwek_2025_39655805),
    get_features_filtered(ref$inhouse)
  )
)

mer <- allmer[,grepl("cytotoxic|gzmb|gzmk", allmer@meta.data[[author_annotation]], ignore.case = T)]
```


```{r}
library(scTypeEval)
DefaultAssay(mer)
data(black_list)
bl <- list(black_list$TCR,
           black_list$Immunoglobulins,
           black_list$Ygenes) |>
       unlist()

gene_list <- list(common_feats = common_feats_ct)

sceval <- create_scTypeEval(allmer,
                            black_list = bl,
                            gene_lists = gene_list)
add_extra_gl <- T

sceval <- run_processing_data(
  scTypeEval = sceval,
  ident = author_annotation,        
  sample = sample_id,  
  min_samples = 1, 
  min_cells = 4
)

if(add_extra_gl){
  sceval <- run_hvg(
    scTypeEval = sceval,
    var_method = "scran",
    ngenes = 2000,
    sample = TRUE, 
    ncores = 6 
  )
  
  # Identify marker genes per cell type
  sceval <- run_gene_markers(
    scTypeEval = sceval,
    method = "scran.findMarkers",
    ngenes_celltype = 100,
    ncores = 6    
  )
}

diss <- c("recip_classif:Match", "Pseudobulk:Cosine",
          "Pseudobulk:Pearson")
for(d in diss){
  sceval <- run_dissimilarity(sceval,
                       method = d,
                       reduction = F,
                       ncores = 6,
                       gene_list = "common_feats")
}
```

```{r}
plot_heatmap(sceval,
             dissimilarity_slot = "recip_classif:Match",
             sort_similarity = "Pseudobulk:Cosine")
```



## custom cosine heatmap

```{r}
d <- c("Pseudobulk:Pearson","Pseudobulk:Cosine")[1]
mat <- sceval@dissimilarity[[d]]@dissimilarity %>% as.matrix()
# get back cosine similarity
mat <- 1 - mat

mph <- data.frame(id = colnames(mat)) %>% 
  separate(id, sep = "_", into = c("Sample_ID", "cell_types"), remove = F) %>% 
  left_join(.,
            sceval@metadata %>%
              distinct(Sample_ID, dataset) %>% 
              mutate(Sample_ID = scTypeEval:::purge_label(Sample_ID)),
            by = "Sample_ID") %>% 
  tibble::column_to_rownames("id") %>% 
  select(-Sample_ID) %>% 
  arrange(dataset, cell_types)

mat <- mat[rownames(mph),rownames(mph)]
phc <- pheatmap(mat = mat,
                cluster_rows = F,
                cluster_cols = F,
                show_rownames = F,
                show_colnames = F,
                annotation_row = mph,
                main = "Pearson correlation - Common genes across datasets")
```

```{r}
feats <- sceval@gene_lists$scran.findMarkers
mat <- sceval@data$pseudobulk@matrix[feats,]

phc <- pheatmap(mat = mat,
                scale = "row",
                cluster_rows = T,
                cluster_cols = F,
                show_rownames = T,
                show_colnames = F,
                annotation_col = mph,
                main = "Cell type markers expression\nPseudobulk per sample")
```



# 7. MetaNeighbor

```{r}
library(MetaNeighbor)
mer0 <- allmer
DefaultAssay(mer0) <- "RNA"
mer0[["HTG"]] <- NULL
mer0[["integrated"]] <- NULL
sce <- as.SingleCellExperiment(mer0)

global_hvgs = variableGenes(dat = sce,
                            exp_labels = sce$dataset)
length(global_hvgs)
```
## Auroc heatmap
```{r}
aurocs <- MetaNeighborUS(var_genes = global_hvgs,
                         dat = sce,
                         study_id = sce$dataset,
                         cell_type = sce$cell_types,
                         fast_version = TRUE, 
                         symmetric_output = TRUE)
```

```{r}
plotHeatmap(aurocs, cex = 0.5)
```

```{r}
new_fig <- "~/OneDrive - Université de Genève/CSI/LabPapers/Cenerenti_KLRG1_CTL/figures/new_figs_Sep26"

fig_path <- file.path(new_fig, "MetaNeighbor_heatmap_otherDatasets.pdf")

pdf(fig_path, width = 6, height = 5.5)
plotHeatmap(aurocs, cex = 0.5)
dev.off()

```



## tophits

```{r}
top_hits <- topHits(aurocs,
                    dat = sce,
                    study_id = sce$dataset,
                    cell_type = sce$cell_types,
                    threshold = 0.5)

top_hit_study <- topHitsByStudy(
  aurocs,
  threshold = 0.8,
  n_digits = 2,
  collapse_duplicates = TRUE
)
```

## Best-hits

```{r}
auroc_bh <- MetaNeighborUS(var_genes = global_hvgs,
                           dat = sce,
                           study_id = sce$dataset,
                           cell_type = sce$cell_types,
                           fast_version = TRUE, 
                           symmetric_output = TRUE,
                           one_vs_best = T)
```

```{r}
plotHeatmap(auroc_bh, cex = 0.5)
```








# to do:

- Instead of only comparing expression of individual markers, define a cytotoxic gene set and calculate a module score in all datasets.

- Compare the populations using correlation / transcriptional similarity: calculate pairwise Spearman correlations, of pseudoubulk per sample. Use sctypeeval? all genes?

- Overlap of GSEA ??


