---
title: "Cross-dataset CTL CD4+ T cells comparison"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Context
In JEM revision, R3 in point 2 pointed:

*Many prior studies have also identified CD4 T cells with cytotoxic features (e.g. PMID: 41043413, PMID: 40299576, PMID: 39655805, and others). It would be valuable if the authors could compare the population identified in this study with previously described cytotoxic CD4 T cell subsets in cancer and other diseases, particularly in terms of their phenotypic and functional overlap.*

We tried to fetch these datasets but either data was not accessible or authors did not share annotation. We could only collect 2 datasets:

### 40299576
   - Actually proposed by reviewer
   - Paper: https://insight.jci.org/articles/view/177373
   Identification and regulation of circulating tumor-TCR-matched cytotoxic CD4+ lymphocytes by KLRG1 in bladder cancer
   - Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE293860
   * Processed (cached): `2.3.Oh_processing.Rmd`
   
### 32497499
Proposed an extra dataset by Camila
Intratumoral CD4+ T Cells Mediate Anti-tumor Cytotoxicity in Human Bladder Cancer
https://doi.org/10.1016/j.cell.2020.05.017
- Data source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149652
* Processed (cached): `2.4.Kwek_processing.Rmd`


# Goals
- Signature overlap and label-transfer comparison against these published cytotoxic CD4 datasets.
- Add heatmap of conserved and unique markers.


# 0 - Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  echo = TRUE
)
```


## Reproducible R environment (recommended)

To reproduce the exact software environment used for this analysis,
restore packages from the repository `renv.lock` file before running
the pipeline.

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

## Load and attach packages

```{r load_packages, message=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(STACAS)
library(SignatuR)
library(BiocParallel)
library(UCell)
```

# 1. Load processed datasets

These were harmonized, QC filtered and normalized in `2.3.Oh_processing.Rmd` and `2.4.Kwek_processing.Rmd`, respectively. 

In both, author's annotation are in `cell_types` and sample_id in `Sample_ID`.

```{r ident_sampleid}
author_annotation <- "cell_type"
sample_id <- "Sample_ID"
```

Load Seurat into a list to run all code in loops
```{r load_seurat}
file_path <- c("cache/Oh_2020_32497499.rds",
               "cache/Kwek_2025_39655805.rds")
seu_list <- bplapply(file_path,
                     BPPARAM = MulticoreParam(workers = 2, progressbar = T),
                     function(f){
                        readRDS(f)
                     })
```



