---
title: "GeoMx Spatial — 1. Data Processing"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

This notebook imports GeoMx/NanoString spatial expression tables and saves a single cached object as an `.rds`.

The raw input files are hosted on Zenodo (DOI: <https://doi.org/10.5281/zenodo.18761515>) and are downloaded automatically into `GeoMx_spatial/cache/`.
For the CD4/CD20 segments used here, the relevant archive in the record is `data_CD4CD20_segments.zip`.

**Output:** `GeoMx_spatial/cache/GeoMx_rawdata_metadata.rds` — a list with:

- `$raw.data`: raw expression matrix (genes × spots)
- `$meta.data`: spot metadata (one row per spot)

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(tidyverse)
library(openxlsx)
library(here)
library(Matrix)
```

```{r paths}
cache_dir <- here::here("GeoMx_spatial", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

# GeoMx input files are downloaded from Zenodo into `cache_dir`.
zenodo_record <- "18761515"  # DOI: 10.5281/zenodo.18761515
```

---

# 1 — Download data from Zenodo

```{r download}
base_url <- paste0("https://zenodo.org/records/", zenodo_record, "/files")

# Zenodo archive containing the CD4/CD20 segments tables.
zip_name <- "data_CD4CD20_segments.zip"
zip_path <- file.path(cache_dir, zip_name)

if (!file.exists(zip_path)) {
  url <- paste0(base_url, "/", zip_name, "?download=1")
  cat("Downloading ", zip_name, " ...\n", sep = "")
  utils::download.file(url = url, destfile = zip_path, mode = "wb", quiet = FALSE)
} else {
  cat(zip_name, " already present — skipping download.\n", sep = "")
}

# Unzip (idempotent-ish): only unzip if the expected folder isn't present.
unzipped_dir <- file.path(cache_dir, "data_CD4CD20 segments")
if (!dir.exists(unzipped_dir)) {
  cat("Unzipping ", zip_name, " ...\n", sep = "")
  utils::unzip(zipfile = zip_path, exdir = cache_dir)
} else {
  cat("Zip already extracted — skipping unzip.\n")
}

# Input files inside the extracted folder
annotation_file <- file.path(unzipped_dir, "Annotation.xlsx")
raw_file <- file.path(unzipped_dir, "Raw data.xlsx")
probe_file <- file.path(unzipped_dir, "Annotation_Probe.xlsx")

missing <- c(annotation_file, raw_file, probe_file)[!file.exists(c(annotation_file, raw_file, probe_file))]
if (length(missing) > 0) {
  stop("Missing expected file(s) after unzip:\n", paste("- ", missing, collapse = "\n"))
}
```

---

# 2 — Import GeoMx tables

This workflow imports the **CD4/CD20 segments** (lymphoid) and builds a single raw counts matrix plus metadata.

## CD4/CD20 segments

```{r import-lymphoid}
annot_lym <- openxlsx::read.xlsx(annotation_file)

annot_lym <- annot_lym %>%
  mutate(
    Patient_letter = Patient,
    Patient = factor(
      Patient,
      levels = c("A", "B", "C", "D", "E"),
      labels = c("LAU706", "LAU380", "LAU1417", "LAU372", "LAU1397")
    )
  )

raw_lym <- openxlsx::read.xlsx(raw_file)
```

## Probe → gene mapping (if needed)

In the original dataset, some raw tables use probe-coded gene identifiers. If your raw tables already use gene symbols as rownames/first column, you can skip this.

```{r import-probes}
probe_lym <- openxlsx::read.xlsx(probe_file) %>%
  filter(CodeClass == "Endogenous")

probe_map <- probe_lym %>%
  select(TargetName, DisplayName)
```

---

# 2 — Build `$raw.data` and `$meta.data`

This block standardizes the raw tables into a genes × spots numeric matrix and concatenates spot metadata.

```{r build-geomx-object}
# Helper: convert a GeoMx raw table to a numeric matrix.
# Assumption: first column contains gene/probe IDs; remaining columns are spots.
raw_to_matrix <- function(raw_df) {
  raw_df <- as.data.frame(raw_df)
  gene_col <- raw_df[[1]]
  mat <- raw_df[, -1, drop = FALSE]
  mat <- as.matrix(suppressWarnings(apply(mat, 2, as.numeric)))
  rownames(mat) <- make.unique(as.character(gene_col))
  mat
}

mat_lym <- raw_to_matrix(raw_lym)

# Optional: map probe IDs to gene symbols when they match TargetName
map_rows_to_gene <- function(mat, probe_map) {
  if (!all(c("TargetName", "DisplayName") %in% colnames(probe_map))) return(mat)
  idx <- match(rownames(mat), probe_map$TargetName)
  new_names <- rownames(mat)
  new_names[!is.na(idx)] <- probe_map$DisplayName[idx[!is.na(idx)]]
  rownames(mat) <- make.unique(new_names)
  mat
}

mat_lym <- map_rows_to_gene(mat_lym, probe_map)

raw_mat <- mat_lym

# Build metadata. Ensure rownames correspond to spot IDs == colnames(raw_mat)
meta_lym <- annot_lym

# Try to detect spot ID column
spot_col <- function(df) {
  candidates <- c("SegmentDisplayName", "Segment", "SegmentID", "ROI", "Spot", "Sample_ID")
  hit <- candidates[candidates %in% colnames(df)][1]
  if (!is.na(hit)) return(hit)
  # fallback: first column
  colnames(df)[1]
}

spot_lym <- spot_col(meta_lym)

meta <- meta_lym %>%
  mutate(Spot = as.character(.data[[spot_lym]])) %>%
  distinct(Spot, .keep_all = TRUE)

# Align metadata row order to matrix columns
common_spots <- intersect(colnames(raw_mat), meta$Spot)
raw_mat <- raw_mat[, common_spots, drop = FALSE]
meta <- meta %>% filter(Spot %in% common_spots) %>% arrange(match(Spot, common_spots))
rownames(meta) <- meta$Spot

geomx <- list(
  raw.data = raw_mat,
  meta.data = meta
)

str(geomx, max.level = 1)
```

---

# 3 — Save cache object

```{r save-cache}
out_path <- file.path(cache_dir, "GeoMx_rawdata_metadata.rds")
saveRDS(geomx, out_path)
cat("Saved:", out_path, "\n")
```

---

# Session information

```{r session-info}
sessionInfo()
```
