---
title: "GeoMx Spatial — 2. DEG Analysis"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

Differential expression analysis on GeoMx spatial spots (CD4 spots) using DESeq2 normalization/variance stabilization.


**Input:** `GeoMx_spatial/cache/GeoMx_raw.data_metadata.rds`

This notebook generates DEG results and then produces volcano plots directly from those results.

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(here)
library(tidyverse)
library(DESeq2)
library(ggrepel)
library(ggpubr)
library(scales)
library(patchwork)
```

```{r parameters}
# Volcano thresholds (match paper defaults if needed)
fdr <- 0.05
fc <- 1.5
```

```{r paths}
cache_dir <- here::here("GeoMx_spatial", "cache")
utils_dir <- here::here("GeoMx_spatial", "utils")

geomx_path <- file.path(cache_dir, "GeoMx_rawdata_metadata.rds")
volcano_util_path <- file.path(utils_dir, "volcano_utils.R")
```

---

# 1 — Load GeoMx cache

```{r load-geomx}
if (!file.exists(geomx_path)) {
  warning(
    "Missing GeoMx cache: ", geomx_path,
    "\nRun GeoMx_spatial/1.Data_processing.Rmd first (chunks: `download`, then `save-cache`).",
    call. = FALSE
  )
  stop("Cannot proceed without GeoMx cache.", call. = FALSE)
}
geomx <- readRDS(geomx_path)

sample_info <- geomx$metadata %>% 
  filter(Type != "perilipin")
raw.data <- geomx$rawdata[,rownames(sample_info)]

stopifnot(!is.null(raw.data), !is.null(sample_info))

cat("Raw matrix:", nrow(raw.data), "genes x", ncol(sample_info), "spots\n")
cat("Metadata:", nrow(sample_info), "spots x", ncol(sample_info), "columns\n")

if (!file.exists(volcano_util_path)) {
  stop("Missing volcano utility: ", volcano_util_path, call. = FALSE)
}
source(volcano_util_path)
```

---

## 1A - summarize table

Summarize spots by type, distance to tumor and B cells and patient of origin.

Corresponding to Suppl Figure 4A

```{r summarize-table}

tab_spots <- sample_info %>%
  group_by(Patient) %>%
  summarise(
    CD4_inside  = sum(Type == "CD4"  & Localisation == "inside"),
    CD4_outside = sum(Type == "CD4"  & Localisation == "outside"),
    CD4_closeB  = sum(Type == "CD4"  & B_dist == "close", na.rm = TRUE),
    CD4_farB    = sum(Type == "CD4"  & B_dist == "far", na.rm = TRUE),
    CD20_inside = sum(Type == "CD20" & Localisation == "inside"),
    CD20_outside= sum(Type == "CD20" & Localisation == "outside"),
    Total_spots = n()
  ) %>%
  arrange(Patient)

tab_spots
```

---

# 2 — DEG comparisons (match GeoMx_NGS_analysis structure)


- Filter to CD4 spots with non-missing `B_dist`
- DESeq2 size factor normalization
- Variance stabilizing transform (`vst`)
- Run DESeq and compute shrunken LFC (`apeglm`). This section mirrors the two comparison types used in `GeoMx_NGS_analysis.Rmd`:

- **Tumor distance:** `Localisation` (outside vs inside) for **CD20+** and **CD4+** spots → **Fig 4D** (left/right)
- **CD4 distance to CD20:** `B_dist` (far vs close) within **CD4+ inside** and **CD4+ outside** → **Fig 4E** (left/right)

```{r deg-helpers}
run_deseq2_apeglm <- function(count_mat, coldata, design_formula) {
  dds <- DESeqDataSetFromMatrix(
    countData = round(count_mat),
    colData = coldata,
    design = design_formula
  )
  dds <- dds[rowSums(counts(dds)) > 10, ]
  dds2 <- DESeq(dds)
  res <- lfcShrink(dds2, coef = resultsNames(dds2)[2], type = "apeglm") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene") %>%
    mutate(padj = ifelse(is.na(padj), 1, padj))
  res
}
```

---

# 2A — Tumor distance (outside vs inside)

Corresponding to **Fig 4D (left): CD20+ spots — outside vs inside**

```{r fig4d-left-cd20-localisation}
imm_cd20 <- sample_info %>%
  filter(Type == "CD20") %>%
  filter(Localisation %in% c("inside", "outside")) %>%
  drop_na(Localisation)

count_cd20 <- raw.data[, rownames(imm_cd20), drop = FALSE]
col_cd20 <- imm_cd20 %>%
  select(Localisation, Patient, Type)

res_cd20_loc <- run_deseq2_apeglm(count_cd20, col_cd20, ~ Localisation)

volcano_plot(res_df = res_cd20_loc)

```

Corresponding to **Fig 4D (right): CD4+ spots — outside vs inside**

```{r fig4d-right-cd4-localisation}
imm_cd4 <- sample_info %>%
  filter(Type == "CD4") %>%
  filter(Localisation %in% c("inside", "outside")) %>%
  drop_na(Localisation)

count_cd4 <- raw.data[, rownames(imm_cd4), drop = FALSE]
col_cd4 <- imm_cd4 %>%
  select(Localisation, Patient, Type, B_dist)

res_cd4_loc <- run_deseq2_apeglm(count_cd4, col_cd4, ~ Localisation)

volcano_plot(res_df = res_cd4_loc)
```

---

# 2B — CD4 distance to CD20 (far vs close)

Corresponding to **Fig 4E (left): CD4+ spots inside tumor — far vs close**

```{r fig4e-left-cd4-bdist-inside}
if (!"B_dist" %in% colnames(sample_info)) {
  message("Skipping B_dist comparison: `B_dist` column not present.")
} else {
  imm_cd4_inside <- sample_info %>%
    filter(Type == "CD4", Localisation == "inside") %>%
    filter(B_dist %in% c("close", "far")) %>%
    drop_na(B_dist)

  count_cd4_inside <- raw.data[, rownames(imm_cd4_inside), drop = FALSE]
  col_cd4_inside <- imm_cd4_inside %>%
    select(B_dist, Patient, Type, Localisation)

  res_cd4_bdist_inside <- run_deseq2_apeglm(count_cd4_inside, col_cd4_inside, ~ B_dist)

  volcano_plot(res_df = res_cd4_bdist_inside)
}
```

Corresponding to **Fig 4E (right): CD4+ spots outside tumor — far vs close**

```{r fig4e-right-cd4-bdist-outside}
if (!"B_dist" %in% colnames(sample_info)) {
  message("Skipping B_dist comparison: `B_dist` column not present.")
} else {
  imm_cd4_outside <- sample_info %>%
    filter(Type == "CD4", Localisation == "outside") %>%
    filter(B_dist %in% c("close", "far")) %>%
    drop_na(B_dist)

  count_cd4_outside <- raw.data[, rownames(imm_cd4_outside), drop = FALSE]
  col_cd4_outside <- imm_cd4_outside %>%
    select(B_dist, Patient, Type, Localisation)

  res_cd4_bdist_outside <- run_deseq2_apeglm(count_cd4_outside, col_cd4_outside, ~ B_dist)

  volcano_plot(res_df = res_cd4_bdist_outside)
}
```

---


# Session information

```{r session-info}
sessionInfo()
```
