---
title: "GeoMx Spatial — 3. Heatmap markers"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

Reproduce the **TFH markers heatmap** corresponding to Fig 4F.
GeoMx cache produced in `GeoMx_spatial/1.Data_processing.Rmd`.

**Input:** `GeoMx_spatial/cache/GeoMx_rawdata_metadata.rds`

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(here)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
```

```{r paths}
cache_dir <- here::here("GeoMx_spatial", "cache")
geomx_path <- file.path(cache_dir, "GeoMx_rawdata_metadata.rds")
```

---

# 1 — Load GeoMx cache

```{r load-geomx}
if (!file.exists(geomx_path)) {
  warning(
    "Missing GeoMx cache: ", geomx_path,
    "\nRun GeoMx_spatial/1.Data_processing.Rmd first (chunks: `download`, then `save-cache`).",
    call. = FALSE
  )
  stop("Cannot proceed without GeoMx cache.", call. = FALSE)
}
geomx <- readRDS(geomx_path)

sample_info <- geomx$metadata %>%
  filter(Type != "perilipin")
rawdata <- geomx$rawdata[, rownames(sample_info), drop = FALSE]

stopifnot(!is.null(rawdata), !is.null(sample_info))

cat("Raw matrix:", nrow(rawdata), "genes x", ncol(rawdata), "spots\n")
cat("Metadata:", nrow(sample_info), "spots x", ncol(sample_info), "columns\n")
```

---

# 2 — Heatmap TFH markers

## 2A — Normalize data (VST)

Corresponding to **Fig 4F** heatmap panel (TFH markers).

```{r normalize-vst}
# Filter for CD4 spots with B_dist info
sample_info_cd4 <- sample_info %>%
  filter(Type == "CD4", !is.na(B_dist)) %>%
  arrange(Localisation, B_dist)

countdata <- rawdata[, rownames(sample_info_cd4), drop = FALSE]

dds <- DESeqDataSetFromMatrix(
  countData = countdata,
  colData = sample_info_cd4,
  design = ~ B_dist
)

# Remove genes with very low counts
dds <- dds[rowSums(counts(dds)) > 10, ]

# DESeq2 size factor normalization + VST
DESeq.ds_f <- estimateSizeFactors(dds)
vsd <- vst(DESeq.ds_f, blind = TRUE)

norm_counts <- assay(vsd)
```

## 2B — Select TFH genes via DEG filter

```{r tfh-deg-filter}
# TFH gene set (copied from simple_new_analysis.Rmd)
tfh.genes <- c(
  "BCL6", "CXCR5", "PDCD1", "TOX", "TOX2", "IRF4", "MAF",
  "CBA2T3", "ICOS", "IL6", "IL10", "IL2", "CD28", "TNFRSF4",
  "IL2RA", "CD69", "FAS", "CD40LG", "B3GAT1", "SH2D1A",
  "IL21", "STAT4", "CEBPA", "NFIA", "IL4", "CXCR4", "ASCL2", "S1PR2"
)

dds2 <- DESeq(dds)
resLFC <- lfcShrink(dds2, coef = resultsNames(dds2)[2], type = "apeglm") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  mutate(padj = ifelse(is.na(padj), 1, padj))

tfh_deg <- resLFC %>%
  filter(
    gene %in% tfh.genes,
    padj < 0.6,
    abs(log2FoldChange) > 0.15
  ) %>%
  arrange(log2FoldChange) %>%
  pull(gene)

tfh_deg
```

## 2C — Heatmap

Corresponding to Figure 4F

```{r heatmap-tfh-markers, fig.height=3}

# Column annotations
colmeta <- sample_info_cd4[, c("B_dist"), drop = FALSE] %>% 
  arrange(B_dist)


mat <- norm_counts[tfh_deg, rownames(colmeta), drop = FALSE]

# Row-wise z-score and cap values to match a fixed scale
scale_top <- 1.5
mat_z <- t(scale(t(mat)))
mat_z[mat_z > scale_top] <- scale_top
mat_z[mat_z < -scale_top] <- -scale_top

hm_breaks <- seq(-scale_top, scale_top, length.out = 101)
hm_cols <- colorRampPalette(
  rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu"))
)(length(hm_breaks) - 1)


pheatmap::pheatmap(
  mat = mat_z,
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = colmeta,
  show_colnames = FALSE,
  breaks = hm_breaks,
  color = hm_cols
)
```

---

# Session information

```{r session-info}
sessionInfo()
```
