---
title: "GeoMx Spatial — CD4 Spots Deconvolution"
author: "Josep Garnica <josep.garnicacaparros@unige.ch>"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    code_folding: show
---

# Overview

Computes cell-type signature scores in GeoMx CD4 spots using marker genes derived from single-cell data, then visualizes the scores by spatial metadata.

Notes:
- This notebook requires two inputs:
  - The GeoMx cache produced by `GeoMx_spatial/1.Data_processing.Rmd` (saved as `GeoMx_spatial/cache/GeoMx_rawdata_metadata.rds`).
  - The integrated Seurat object `int` produced in `scRNA-seq/1.1.Data_processing.Rmd` + `scRNA-seq/1.2.Integration_annotation.Rmd` (saved as `scRNA-seq/cache/cd4_total_integrated_manual.rds`).

---

# 0 — Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
```

## Reproducible R environment (recommended)

```{r renv-restore, eval=FALSE}
# Run once from the repository root
install.packages("renv")
renv::restore(prompt = FALSE)
```

```{r packages}
library(here)
library(tidyverse)
library(Seurat)
# Signature scoring
library(UCell)
library(ggpubr)
```

```{r paths}
geomx_cache <- here::here("GeoMx_spatial", "cache", "GeoMx_rawdata_metadata.rds")
sc_cache <- here::here("scRNA-seq", "cache")
int_path <- file.path(sc_cache, "cd4_total_integrated_manual.rds")
```

---

# 1 — Load GeoMx cache and scRNA-seq object

```{r load-inputs}
if (!file.exists(geomx_cache)) {
  stop("Missing GeoMx cache: ", geomx_cache, "\nRun GeoMx_spatial/1.Data_processing.Rmd first.")
}
geomx <- readRDS(geomx_cache)
rawdata <- geomx$rawdata
metadata <- geomx$metadata
stopifnot(!is.null(rawdata), !is.null(metadata))

if (!file.exists(int_path)) {
  stop(
    "Missing integrated scRNA-seq object: ", int_path,
    "\nRun scRNA-seq/1.2.Integration_annotation.Rmd first."
  )
}
int <- readRDS(int_path)
```

---

# 2 — Derive markers from scRNA-seq (TILN only)

This mirrors the reference workflow:

- Use `FindAllMarkers()` on `int` restricted to `tissue == "TILN"` to match the spatial data
- Keep strong positive markers per cluster
- Select up to 10 marker genes per cluster

```{r derive-markers}
Idents(int) <- "annotation"

if (!"tissue" %in% colnames(int@meta.data)) {
  stop("Expected 'tissue' column missing in int@meta.data")
}

de_all0 <- FindAllMarkers(
  int[, int$tissue == "TILN"],
  assay = "RNA",
  logfc.threshold = 0.25,
  test.use = "bimod",
  only.pos = TRUE,
  min.pct = 0.4,
  min.diff.pct = 0.1
)

de_all <- de_all0 %>%
  filter(
    p_val_adj < 0.05,
    avg_log2FC > 0.4,
    !grepl("^HLA|^RPS", gene) # remove HLA and ribosomal genes
  ) %>%
  mutate(
    dif.pct = pct.1 - pct.2,
    ppct = dif.pct / pct.1
  ) %>%
  arrange(cluster, desc(dif.pct))

# Choose best markers for each cell type
split_markers <- split(de_all, de_all$cluster)
for (cl in names(split_markers)) {
  split_markers[[cl]] <- split_markers[[cl]] %>%
    arrange(desc(ppct)) %>%
    filter(!grepl("^MIR", gene))

  if (nrow(split_markers[[cl]]) > 10) {
    split_markers[[cl]] <- split_markers[[cl]][1:10, ]
  }

  split_markers[[cl]] <- split_markers[[cl]] %>% pull(gene)
}

# Show marker set sizes
tibble(
  celltype = names(split_markers),
  n_genes = vapply(split_markers, length, integer(1))
)
```

---

# 3 — Score signatures in GeoMx spots (UCell)

```{r score-ucell}
# rawdata is genes x spots
u_scores <- ScoreSignatures_UCell(
  matrix = rawdata,
  features = split_markers,
  ncores = 4,
  name = ""
)

u_scores <- u_scores %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Spot") %>%
  pivot_longer(-Spot, names_to = "Celltype", values_to = "Score")
```

---

# 4 — Plot signature scores by GeoMx metadata

Corresponding to Figure 4G

```{r deconvolution-boxplot}
if (!all(c("Type", "B_dist") %in% colnames(metadata))) {
  stop("Expected metadata columns missing: need at least Type and B_dist")
}

res_u <- metadata %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Spot") %>%
  right_join(u_scores, by = "Spot") %>%
  filter(
    Type == "CD4",
    !is.na(B_dist)
  )

# Match reference example: Cytotoxic + TFH
res_u2 <- res_u %>%
  filter(Celltype %in% c("Cytotoxic", "TFH"))

the <- theme(
  legend.title = element_text(size = 16, face = "bold"),
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
)

bdist <- res_u2 %>%
  mutate(B_dist = factor(B_dist, levels = c("close", "far"))) %>%
  ggplot(aes(Celltype, Score)) +
  geom_point(
    aes(color = Patient, group = B_dist),
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 1),
    alpha = 0.5,
    show.legend = TRUE
  ) +
  geom_boxplot(
    aes(fill = B_dist),
    color = "black",
    alpha = 0.3,
    outlier.shape = NA,
    position = position_dodge(1)
  ) +
  scale_fill_manual(values = c(
    "close" = "#6C8EBF",
    "far" = "#B07C59"
  )) +
  ggpubr::stat_compare_means(
    aes(group = B_dist),
    method = "wilcox.test",
    label = "p.format",
    p.adjust.method = "BH",
    size = 3,
    tip.length = 0,
    bracket.size = 0.8
  ) +
  labs(fill = "Distance to B cells", y = "Signature Score") +
  facet_wrap(~Localisation) +
  ggpubr::theme_classic2() +
  the

bdist
```

---

# Session information

```{r session-info}
sessionInfo()
```
